\section{Simplicial Homology} \label{sectSimplHomol}

As mentioned in the introduction of this chapter, readers may choose
to only browse through this and the next sections if they are not
very interested in learning about simplicial homology and cohomology.
There are however some concepts and definitions in this two sections
that the reader should learn because they are used in the next
chapter.  In particular, we are talking about the definitions of
simplex and simplicial complex in Section~\ref{subsOShomo}, oriented
simplex in Section~\ref{ssectSimpCo}, and smoothly triangulated
manifold in Section~\ref{ssectSCandCM}.  Sections~\ref{sectSingHom} and
\ref{sectSingCohom} on singular homology and cohomology are the
important sections in this chapter.  In particular,
Subsection~\ref{ssectdeRhamSing} contains the statement and the proof 
of de Rham theorem, Theorem~\ref{deRhamThmSing}. which is the
motivation for this chapter.

We present and sketch the proof of a version of de Rham theorem for
simplicial cohomology in Section~\ref{ssectSCandCM} for those who
may only be interested in a brief study of simplicial homology and
cohomology.

Our presentation of the simplicial homology and cohomology is more for
a pedagogical purpose, to prepare the reader to the study of singular
homology and cohomology.  There is a lot of similarity between the
results and techniques of proof used in simplicial theory and those used in
singularity theory.  This is not surprising as it is explained in
Section~\ref{sectRelSandS}.  Nevertheless, there are really
important and interesting results that can only be deduced from
simplicial homology.  Theorem~\ref{thmEulerAlpha} is one of them.
Both simplicial and singular theories are important.  It is not a
question of which one is better than the other.

\subsection{Simplicial Complexes} \label{ssectSimpCo}

\begin{defn}
A subset $\{\VEC{x}_0,\VEC{x}_1,\VEC{x}_2,\ldots, \VEC{x}_k\}$ of
$\displaystyle \RR^n$ is
{\bfseries convex independent}\index{Convex Independent} if
$\{\VEC{x}_1-\VEC{x}_0,\VEC{x}_2-\VEC{x}_0,\ldots, \VEC{x}_k-\VEC{x}_0\}$
are linearly independent.
\end{defn}

\begin{defn}
Let $s = \{\VEC{x}_0,\VEC{x}_1,\VEC{x}_2,\ldots, \VEC{x}_k\}$ be a
convex independent subset of $\displaystyle \RR^n$.   The set
$\displaystyle \left\{ \sum_{j=0}^k a_j\VEC{x}_j : a_j \geq 0 \ \text{and} \
\sum_{j=0}^k a_k = 1 \right\}$, denoted $[s]$ or
$[\VEC{x}_0,\VEC{x}_1,\VEC{x}_2,\ldots,\VEC{x}_k]$, is called a
{\bfseries closed $\mathbf{k}$-simplex}\index{Closed $k$-Simplex}.
The {\bfseries vertices}\index{Vertex} of $[s]$ are the
points $\VEC{x}_j$ for $0\leq j \leq k$. The
{\bfseries closed faces}\index{Closed Face} of $[s]$ are the
closed $\mathbf{m}$-simplices
$[\VEC{x}_{j_0},\VEC{x}_{j_1},\VEC{x}_{j_2},\ldots, \VEC{x}_{j_m}]$
for $\{ j_0, j_1, \ldots, j_m\} \subset \{0,1,2,\ldots, k\}$
and $0\leq m \leq k$.
The {\bfseries dimension}\index{Dimension} of $[s]$, denoted $\dim \, [s]$,
is $k$.
\end{defn}

The reader should prove that $[s]$ is the smallest
convex set containing $s$.

\begin{defn}
The coefficients $a_j$ in the expression
$\displaystyle \VEC{x} = \sum_{j=0}^k a_j\VEC{x}_j$ with $a_j \geq 0$ and
$\displaystyle \sum_{j=0}^k a_k = 1$ are called
{\bfseries barycentric coordinates}\index{Barycentric Coordinates}
of $x$.
\end{defn}

\begin{defn}
Let $s = \{\VEC{x}_0,\VEC{x}_1,\VEC{x}_2,\ldots, \VEC{x}_k\}$ be a
convex independent subset of $\displaystyle \RR^n$.
The set $\displaystyle \left\{ \sum_{j=0}^k a_j\VEC{x}_j : a_j > 0
\ \text{and} \ \sum_{j=0}^k a_k = 1 \right\}$, denoted $(s)$ or
$(\VEC{x}_0,\VEC{x}_1,\VEC{x}_2, \ldots, \VEC{x}_k)$,
is called an {\bfseries open $\mathbf{k}$-simplex}\index{Open $k$-Simplex}.
The {\bfseries dimension}\index{Dimension} of $(s)$, denoted $\dim\, (s)$,
is $k$.
\end{defn}

\begin{defn}
The {\bfseries open faces}\index{Open Face} of the closed $k$-simplex
$[\VEC{x}_0,\VEC{x}_1,\ldots,\VEC{x}_k]$ are the
open $m$-simplices
$(\VEC{x}_{j_0},\VEC{x}_{j_1},\VEC{x}_{j_2},\ldots, \VEC{x}_{j_m})$
for $\{ j_0, j_1, \ldots, j_m\} \subset \{0,1,2,\ldots, k\}$
and $0\leq m \leq k$.
\end{defn}

A closed $k$-simplex $[\VEC{x}_0,\VEC{x}_1,\ldots,\VEC{x}_k]$ is the
union of all its open faces (Figure~\ref{SimpCpx1}).  The reader must
realize that the closed $0$-complex $[\VEC{x}_j]$ is equal to the open
$0$-complex $(\VEC{x}_j)$.  If $(s_1)$ and $(s_2)$ are two open faces
of a closed $k$-simplex, then either $(s_1) = (s_2)$ or
$(s_1) \cap (s_2) = \emptyset$.

\pdfF{alg_top/simpcpx1}{Closed $3$-simplex}{Illustration of a closed
$3$-simplex
$\displaystyle [\VEC{x}_0,\VEC{x}_1,\VEC{x}_2,\VEC{x}_3] \subset \RR^3$.
This closed simplex is the union of the open faces
$(\VEC{x}_0,\VEC{x}_1,\VEC{x}_2,\VEC{x}_3)$, $(\VEC{x}_0,\VEC{x}_1,\VEC{x}_2)$,
$(\VEC{x}_0,\VEC{x}_1,\VEC{x}_3)$,
$(\VEC{x}_0,\VEC{x}_2,\VEC{x}_3)$, $(\VEC{x}_1,\VEC{x}_2,\VEC{x}_3)$,
$(\VEC{x}_0,\VEC{x}_1)$, $(\VEC{x}_0,\VEC{x}_2)$, $(\VEC{x}_0,\VEC{x}_3)$,
$(\VEC{x}_1,\VEC{x}_2)$, $(\VEC{x}_1,\VEC{x}_3)$, $(\VEC{x}_2,\VEC{x}_3)$,
$(\VEC{x}_0)$, $(\VEC{x}_1)$, $(\VEC{x}_2)$ and $(\VEC{x}_3)$.}{SimpCpx1}

\begin{defn}
A {\bfseries simplicial complex}\index{Simplicial Complex} $K$ in
$\displaystyle \RR^n$ is a finite set of open simplices such that:
\begin{enumerate}
\item If $(s)$ is an open simplices in $K$, then all the open faces of
the closed simplex $[s]$ are also in $K$.
\item If $(s_1)$ and $(s_2)$ are two open simplices in $K$, then
either $(s_1) = (s_2)$ or $(s_1) \cap (s_2) = \emptyset$.
\end{enumerate}
The {\bfseries dimension}\index{Dimension} of the simplicial complex
$K$, denoted $\dim K$, is defined as the dimension of the largest open
simplex in $K$.  The union of the open simplices in $K$ is the subset
of $\displaystyle \RR^n$ denoted $[K]$.
The {\bfseries vertices}\index{Vertices} of $K$ are the elements
$\VEC{x} \in [K]$ such that $(\VEC{x}) \in K$; namely, $(\VEC{x})$ is a
$0$-simplex in $K$.
\end{defn}

We have that $\displaystyle [K] = \bigcup_{(s) \in K} (s)
= \bigcup_{(s) \in K} [s]$ and $[K]$ is a compact subsets of
$\displaystyle \RR^n$ because it is a finite union of the compact sets
$[s]$ for $(s) \in K$.

\pdfF{alg_top/simpcpx2}{Simplicial complex}{The simplicial complex
$K$ represented by the set $\displaystyle [K] \subset \RR^3$ in the
figure above is the union of the set of all the open faces of the
closed $3$-simplex $[\VEC{x}_0,\VEC{x}_1,\VEC{x}_2,\VEC{x}_3]$ listed in
Figure~\ref{SimpCpx1} with the set
$\displaystyle \left\{ (\VEC{x}_0,\VEC{x}_4,\VEC{x}_5),
(\VEC{x}_0,\VEC{x}_4), (\VEC{x}_0,\VEC{x}_5), (\VEC{x}_4,\VEC{x}_5),
(\VEC{x}_4), (\VEC{x}_5) \right\}$.  The dimension of $K$ is $3$.}{SimpCpx2}

All the following notions will be very useful to study simplicial
complexes.

\begin{defn}
A subset $L$ of a simplicial complex $K$ is called a
{\bfseries subcomplex}\index{Subcomplex} of $K$ if $L$ is itself a
simplicial complex.
\end{defn}

\begin{defn}
Let $K$ be a simplicial complex of dimension $k$.  Given
$0 \leq r \leq k$, the {\bfseries $\mathbf{r}$-skeleton}\index{$r$-Skeleton} 
of $K$, denoted $\displaystyle K^r$, is the set
$\displaystyle K^r = \left\{ (s) : (s) \in K \ \text{and}\ \dim\, (s)
\leq r \right\}$.
\end{defn}

Since a closed $k$-simplex $[s]$ is the union of all its open faces,
we may treat $s$ as the set of all the open faces of $[s]$; namely,
we may treat $s$ as a simplicial complex.  Note that all the open
faces of $[s]$ are open $m$-simplices with $m \leq \dim\, [s]$.
By extension, the {\bfseries $r$-skeleton}\index{$r$-Skeleton} of $s$ for
$0 \leq r \leq k$, denoted $\displaystyle s^r$, is the set of all open
faces of $[s]$ corresponding to open $m$-simplex with $m \leq r$.
Moreover, $\displaystyle [s^r]$ will denote the union
of all the open simplices in $\displaystyle s^r$.

\begin{defn}
Given $\displaystyle \VEC{x} \in \RR^n$ and
$\displaystyle A \subset \RR^n$, the pair $(\VEC{x},A)$ is in
{\bfseries general position}\index{General Position} if
$\VEC{x} \not\in A$ and, for any two distinct points
$\VEC{a}_1$ and $\VEC{a}_2$ in $A$, we have that
$[\VEC{x},\VEC{a}_1] \cap [\VEC{x},\VEC{a}_2] = \{ \VEC{x}\}$.
If $(\VEC{x},A)$ is in general position, then the {\bfseries cone}\index{Cone}
with base $A$ and vertex $\VEC{x}$, denoted $\VEC{x} \ast A$, is the set
$\displaystyle \VEC{x} \ast A = \bigcup_{\VEC{a} \in A} [\VEC{x},\VEC{a}]$.

Given a simplicial complex $K$, the {\bfseries cone}\index{Cone}
$\VEC{x} \ast K$ is the simplicial
complex consisting of all the open simplices of the form
$(\VEC{x},\VEC{x}_0, \VEC{x}_1, \ldots , \VEC{x}_k)$ with
all their faces for all
$(\VEC{x}_0, \VEC{x}_1, \ldots , \VEC{x}_m) \in K$ with $m \leq \dim K$.
\end{defn}

The closed $k$-simplex
$[s] = [\VEC{x}_0,\VEC{x}_1, \VEC{x}_2, \ldots, \VEC{x}_k]$ can be
expressed as $\displaystyle [s] = \VEC{x} \ast [s^{k-1}]$ for any
$\VEC{x} \in (s)$.

\subsection{Barycentric Subdivisions} \label{ssectBarSubd}

\begin{defn}
The {\bfseries barycentre}\index{Barycentre} of an open $k$-simplex
$(s) = (\VEC{x}_0, \VEC{x}_1, \ldots, \VEC{x}_k)$ is the point
$\VEC{b}_{(s)} \in (s)$ defined by
$\displaystyle \VEC{b}_{(s)} = (k+1)^{-1} \sum_{j=0}^k \VEC{x}_j$.
\end{defn}

\begin{defn}
A {\bfseries subdivision}\index{Subdivision} of a simplicial complex
$K$ is another simplicial complex $L$ such that $[L] = [K]$ and, if
$(s) \in L$, then $(s)$ is a subset of an open simplex in $K$.
\end{defn}

\pdfF{alg_top/simpcpx3}{Subdivision of a simplicial complex}{
(b) and (c) illustrate two possible subdivisions of the simplicial
complex illustrated in (a).  (b) is not a subdivision of the
simplicial complex illustrated in (c) and vice-versa.}{SimpCpx3}

\begin{defn}
A {\bfseries partial order}\index{Partial order} on a simplicial
complex $K$ is defined as follows.  Given $(s_1)$ and $(s_2)$ in $K$,
we write that $(s_1) \leq (s_2)$ if $(s_1)$ is an open face of
$[s_2]$.  If in addition $(s_1) \neq (s_2)$, then we write that
$(s_1) < (s_2)$.
\end{defn}

The following result will play a fundamental role in the ``simplicial
approximation'' of ``simplicial function'' in the next subsection.
Its proof is not too complicated but it is long (see \cite{ST}).  It
uses the concept of general position introduced above.

\begin{theorem}
Suppose that $K$ is a simplicial complex in $\displaystyle \RR^n$.  Let
$\displaystyle K^{[1]}$ be the simplicial complex defined by
\begin{align*}
K^{[1]} & = \left\{ (\VEC{b}_{(s_0)}, \VEC{b}_{(s_1)},\ldots, \VEC{b}_{(s_m)}) :
(s_1) < (s_2) < \ldots < (s_m) \ \text{and} \right. \\
&\hspace{13em} \left.  \ (s_j) \in K \ \text{for}
\ 0 \leq j \leq m \leq \dim K  \right\} \ .
\end{align*}
Then $\displaystyle K^{[1]}$ is a subdivision of $K$
(Figure~\ref{SimpCpx4}).  Moreover, we have that \\
$(\VEC{b}_{(s_0)}, \VEC{b}_{(s_1)},\ldots, \VEC{b}_{(s_m)})$ is a
subset of $(s_m)$ for all sequences
$(s_1) < (s_2) < \ldots < (s_m)$ of open simplices in $K$.
\end{theorem}

\pdfF{alg_top/simpcpx4}{First barycentric subdivision}{We have
illustrated some of the open simplices in the simplicial complex
$\displaystyle s^{[1]}$ where $s$ is the simplicial complex
associated to the closed $2$-simplex $[s] = [x_0,x_1,x_2]$.  We have
drawn in blue the open $2$-simplex
$(\VEC{b}_{(s_1)}, \VEC{b}_{(s_2)}, \VEC{b}_{(s_3)})$ for $(s_1) = (\VEC{x}_1)$,
$(s_2) = (\VEC{x}_1,\VEC{x}_2)$ and $(s_3) = (s)$.  We effectively
have that $(s_1) < (s_2) < (s_3)$.  We have drawn in red the
open $1$-simplex $(\VEC{b}_{(s_4)}, \VEC{b}_{(s_3)})$ for
$(s_4) = (\VEC{x}_2)$ and $(s_3)$.  Again, we have that $(s_4) < (s_3)$.
Note that the open $0$-simplex $(\VEC{b}_{(t)})$ is associated to a
$0$-simplex $(t)$ is $(t)$ itself.}{SimpCpx4}

It follows from the definition of $\displaystyle K^{[1]}$ that
$\displaystyle \dim K^{[1]} = \dim K$.

\begin{defn} \label{defnBarySubd}
Let $K$ be a simplicial complex in $\displaystyle \RR^n$.
The simplicial complex $\displaystyle K^{[1]}$ is called the
{\bfseries first barycentric subdivision}\index{Barycentric
Subdivision!First Barycentric Subdivision} of $K$.  Inductively, we
may define the {\bfseries $\displaystyle \mathbf{m^{th}}$ barycentric
subdivision}\index{Barycentric Subdivision!$m^{th}$
Barycentric Subdivision} of $K$ as the simplicial complex
$\displaystyle K^{[m]} = (K^{[m-1]})^{[1]}$ for $m>1$.
\end{defn}

Recall that the diameter of a set $\displaystyle S \subset \RR^n$ is defined
by $\displaystyle \diam S = \inf_{\VEC{x},\VEC{y} \in S}
\|\VEC{x}- \VEC{y}\|$ where $\|\cdot\|$ is a norm on
$\displaystyle \RR^n$, usually the Euclidean norm. 

\begin{lemma} \label{lemDiams}
If $[s] = [\VEC{x}_0,\VEC{x}_1,\VEC{x}_2,\ldots, \VEC{x}_k]$ is a
closed $k$-simplex, then $\diam\, [s] = \|\VEC{x}_{j_1}  - \VEC{x}_{j_2}\|$
for some $j_1,j_2 \in \{0,1,2,\ldots,k\}$.
\end{lemma}

\begin{proof}
Since $[s]$ is a compact set, there exist $\VEC{y}_1$ and $\VEC{y}_2$
in $[s]$ such that $\diam\, [s] = \|\VEC{y}_1 - \VEC{y}_2\|$.
Without loss of generality, suppose that $\VEC{y}_2$ is not a vertex.
Then we may express $\VEC{y}_2$ as
$\VEC{y}_2 = p \VEC{w}_1 + (1-p) \VEC{w}_2$ for some $0 < p < 1$ and
$\VEC{w}_1, \VEC{w}_2 \in [s]$.  Let
$g(t) = \|\VEC{y}_1 - t \VEC{w}_1 - (1-t) \VEC{w}_2\|$ for
$0 \leq t \leq 1$.  We have that $g$ is a convex function; namely,
$g(tq_1 + (1-t)q_2) \leq t g(q_1) + (1-t) g(q_2)$ for $0\leq q_1,q_2 \leq 1$
and $0 \leq t \leq 1$.  Here is a possible graph of $g$.
\pdfbox{alg_top/convex}
Thus $g$ reaches its maximum at one of the end points.  This implies
that $\diam\, [s] = g(p) < \max\{ g(0),g(1)\}
= \max \{ \|\VEC{y}_1 - \VEC{w}_1\|, \|\VEC{y}_1 -\VEC{w}_2\| \}$. 
This is a contradiction.
\end{proof}

\begin{defn}
Let $K$ be a simplicial complex.  The {\bfseries mesh}\index{Mesh}
of $K$ is defined as $\displaystyle \mesh K = \max_{(s) \in K} \diam\, [s]$.
\end{defn}

\begin{prop} \label{propMeshMBS}
Suppose that $K$ is a simplicial complex of dimension $k$.  Then
$\displaystyle \mesh K^{[1]} \leq (k/(k+1)) \mesh K$.
\end{prop}

\begin{proof}
By definitions of $\displaystyle K^{[1]}$ and the mesh of a simplicial
complex, there exists $(s_1) < (s_2) < \ldots < (s_m)$ in $K$ with $m \leq k$
such that
$\displaystyle \mesh K^{[1]}
= \diam\, [\VEC{b}_{(s_0)}, \VEC{b}_{(s_1)},\ldots, \VEC{b}_{(s_m)}]$.
It follows from Lemma~\ref{lemDiams} that
$\displaystyle \mesh^{[1]} K = \|\VEC{b}_{(s_p)} - \VEC{b}_{(s_q)}\|$ for some
$p,q \in \{0,1,\ldots,m\}$.  Without loss of generality, we may
re-index the vertices of $K$ and assume
that $p < q$, $(s_p) = (\VEC{x}_0,\VEC{x}_1,\ldots, \VEC{x}_p)$ and
$(s_q) = (\VEC{x}_0,\VEC{x}_1,\ldots, \VEC{x}_q)$.  Hence
\begin{align*}
&\mesh K^{[1]} = \|\VEC{b}_{(s_p)} - \VEC{b}_{(s_q)}\|
= \left\| \frac{1}{p+1} \sum_{j=0}^p \VEC{x}_j - 
\frac{1}{q+1} \sum_{i=0}^q \VEC{x}_i \right\| \\
&= \left\| \left(\frac{1}{p+1}- \frac{1}{q+1}\right) \sum_{j=0}^p \VEC{x}_j - 
\frac{1}{q+1} \sum_{i=p+1}^q \VEC{x}_i \right\|
= \left\| \frac{q-p}{(p+1)(q+1)}\sum_{j=0}^p \VEC{x}_j - 
\frac{1}{q+1} \sum_{i=p+1}^q \VEC{x}_k \right\| \\
&= \frac{1}{q+1}\left\| \frac{q-p}{p+1}\sum_{j=0}^p \VEC{x}_j - 
\sum_{i=p+1}^q \VEC{x}_i \right\|
= \frac{1}{q+1}\left\| \frac{q+1}{p+1}\sum_{j=0}^p \VEC{x}_j - 
\sum_{i=0}^q \VEC{x}_i \right\| \\
&= \frac{1}{q+1}\left\| \sum_{i=0}^q \left( \frac{1}{p+1}\sum_{j=0}^p
\VEC{x}_j - \VEC{x}_i \right) \right\|
= \frac{1}{(q+1)(p+1)} \left\| \sum_{i=0}^q \left( \sum_{j=0}^p
(\VEC{x}_j - \VEC{x}_i) \right) \right\| \\
&\leq \frac{1}{(q+1)(p+1)} \sum_{i=0}^q \sum_{j=0}^p
\left\|  \VEC{x}_j - \VEC{x}_i\right\| \ ,
\end{align*}
where we have used $(q-p) = (q+1) - (p+1)$ to get the fifth equality.
There are $(p+1)(q+1) - (p+1) = q(p+1)$ terms in the last sum because
$(p+1)$ of the terms are null, those for $i = j$.  Moreover
$\left\|  \VEC{x}_j - \VEC{x}_i\right\| \leq \mesh K$ for all $i$ and $j$.
thus
\[
\mesh K^{[1]} \leq \frac{q(p+1)}{(q+1)(p+1)} \mesh K 
= \frac{q}{q+1} \mesh K \leq \frac{k}{k+1} \mesh K
\]
because $q \leq k$.
\end{proof}

It follows from the previous proposition that
$\displaystyle \mesh K^{[m]} \leq \big( (k/(k+1)\big)^m \mesh K \to 0$
as $m \to \infty$.

\subsection{Simplicial Approximation} \label{ssectSA}

\begin{defn} \label{defnsimplMaps}
Let $K$ and $L$ be two simplicial complexes.  A
{\bfseries simplicial map}\index{Simplicial Map} between $K$ and $L$ is a
function $\phi :[K] \to [L]$ that satisfies the following conditions.
\begin{enumerate}
\item $\phi(\VEC{x})$ is a vertex of $L$ if $\VEC{x}$ is a vertex of $K$.
\item If $(s) = (\VEC{x}_0,\VEC{x}_1, \ldots, \VEC{x}_k)\in K$, then
there exists $(t) \in L$ such that $\phi(\VEC{x}_i) \in [t]$ for
$0 \leq i \leq k$. 
\item If $\VEC{x} \in (s) = (\VEC{x}_0,\VEC{x}_1, \ldots, \VEC{x}_k)$
and $\displaystyle \VEC{x} = \sum_{j=0}^k a_j \VEC{x}_j$ with $a_j >0$
and $\displaystyle \sum_{j=0}^k a_j = 1$, then 
$\displaystyle \phi(\VEC{x}) = \sum_{j=0}^k a_j \phi(\VEC{x}_j)$.
\end{enumerate}
\end{defn}

The set $\displaystyle \bigg\{ \phi(\VEC{x}) :
\displaystyle \VEC{x} = \sum_{j=0}^k a_j \VEC{x}_j \ \text{with}\ a_j >0
\ \text{and} \ \sum_{j=0}^k a_j = 1\bigg\}$ in (3) represents an
open face of $[t]$ mentioned in (2).  It may not be $(t)$ itself because
$\phi(\VEC{x}_i)$ for $0\leq i \leq k$ may not be all
the vertices of $[t]$.

\begin{defn} \label{defnStarV}
Let $K$ be a simplicial complex and $\VEC{x}$ be a vertex of
$K$.  The {\bfseries star}\index{Star} of $\VEC{x}$, denoted $\St(\VEC{x})$,
is the set defined by $\displaystyle
\St(\VEC{x}) = \bigcup_{\substack{(s) \in K\\\VEC{x} \in [s]}} (s)$
(Figure~\ref{Star1}).
\end{defn}

\pdfF{alg_top/star1}{Star of a vertex}{We have illustrated on the
left a simplicial complex $K$ where $[K]$ is the union of the
closed simplices $[\VEC{x}_0,\VEC{x}_1,\VEC{x}_6]$,
$[\VEC{x}_1,\VEC{x}_5,\VEC{x}_6]$, $[\VEC{x}_3,\VEC{x}_4,\VEC{x}_5]$, 
$[\VEC{x}_1,\VEC{x}_2]$ and $[\VEC{x}_2,\VEC{x}_3]$.  We have
illustrated on the right in blue the star of $\VEC{x}_1$.}{Star1}

Given a vertex $\VEC{x}$ of a simplicial complex $K$, we have that
$\St(\VEC{x})$ is open in the induced topology on $[K]$ from
$\displaystyle \RR^n$.  Moreover, $\VEC{x}$ is the only vertex of $K$ in
$\St(\VEC{x})$ and
$\displaystyle \{\St(\VEC{x}) : \VEC{x} \ \text{is a vertex of} \ K\}$
is an open cover of $[K]$ with respect to the induced topology
on $[K]$ from $\displaystyle \RR^n$.

\begin{defn}
Let $K$ and $L$ be two simplicial complexes.  A
simplicial map $\phi:[K]\to [L]$ is a {\bfseries simplicial
approximation}\index{Simplicial Approximation} of a continuous
function $f:[K] \to [L]$ if
$f( \St(\VEC{x}) ) \subset \St(\phi(\VEC{x}))$ for all vertices
$\VEC{x}$ of $K$.
\end{defn}

\begin{prop} \label{propNfmg}
Suppose that $K$ and $L$ are two simplicial complexes.
If $\phi:[K]\to [L]$ is a simplicial approximation of a continuous
function $f:[K] \to [L]$ and $\VEC{x} \in [K]$, then
$f(\VEC{x})$ and $\phi(\VEC{x})$ belongs to the same closed simple
$[t]$ with $(t) \in L$.
\end{prop}

\begin{proof}
Given $\VEC{x} \in [K]$, there exists an open simplex
$(s) = (\VEC{x}_0,\VEC{x}_1,\ldots, \VEC{x}_m) \in K$ such that
$\VEC{x} \in (s)$.  Since
$\VEC{x} \in (s) \subset \St(\VEC{x}_j)$ for all $0\leq j \leq m$, we
get from the definition of simplicial approximation that
$f(\VEC{x}) \in f(\St(\VEC{x}_j)) \subset \St(\phi(\VEC{x}_j))$
for $0 \leq j \leq m$.

There exists an open simplex $(t) \in L$ such that $f(\VEC{x}) \in (t)$.
Since $f(\VEC{x}) \in (t) \cap \St(\phi(\VEC{x}_j))$ for
$0 \leq j \leq m$, we get that
$(t) \cap \St(\phi(\VEC{x}_j)) \neq \emptyset$
for $0 \leq j \leq m$.  By definition of simplicial complex, we
have that $(t) \subset \St(\phi(\VEC{x}_j))$
for $0 \leq j \leq m$.  By definition of the star of a vertex, we
have that $\phi(\VEC{x}_j)$ for $0 \leq j \leq m$ are vertices of 
$[t]$.  Note that they may not be all the vertices of $[t]$.

Since $\VEC{x} \in (s)$, we have that
$\displaystyle \VEC{x} = \sum_{j=0}^m a_j \VEC{x}_j$ for some
$a_j> 0$ such that $\displaystyle \sum_{j=0}^m a_j = 1$.  Hence, we
get from the definition of simplicial map that
$\displaystyle \phi(\VEC{x}) = \sum_{j=0}^m a_j \phi(\VEC{x}_j)$ with
$a_j> 0$ such that $\displaystyle \sum_{j=0}^m a_j = 1$.  Thus
$\phi(\VEC{x}) \in \big(\phi(\VEC{x}_0), \phi(\VEC{x}_1), \ldots,
\phi(\VEC{x}_k)\big) \subset [t]$.  Note that we may not have that
$\big(\phi(\VEC{x}_0), \phi(\VEC{x}_1), \ldots, \phi(\VEC{x}_k)\big) = (t)$
because, as we said above, $\phi(\VEC{x}_j)$ for $0 \leq j \leq m$ may
not be all the vertices of $[t]$.  Therefore
$\big(\phi(\VEC{x}_0), \phi(\VEC{x}_1), \ldots, \phi(\VEC{x}_k)\big)$
may be only an open face of $[t]$ not equal to $(t)$ itself..
\end{proof}

This proposition implies that
$\displaystyle \|f - \phi\|_{[K]} = \max_{\VEC{x} \in [K]}
\|f(\VEC{x}) - \phi(\VEC{x}\| \leq \mesh L$.

\begin{prop}
Suppose that $K$ and $L$ are two simplicial complexes.
If $\phi:[K]\to [L]$ is a simplicial approximation of a simplicial map
$f:[K] \to [L]$, then $\phi = f$.
\end{prop}

\begin{proof}
For each vertex $\VEC{x} \in K$, we have that
$f(\VEC{x}) \in f(\St(\VEC{x})) \subset \St(\phi(\VEC{x}))$.
Since $f$ is a simplicial map, $f(\VEC{x})$ is a vertex of $L$.
Since $\St(\phi(\VEC{x}))$ contains only one vertex of $L$, namely
$\phi(\VEC{x})$, we get that $f(\VEC{x}) = \phi(\VEC{x})$.

Since $f$ and $\phi$ are simplicial maps that take the same value at
all the vertices of $K$, then it follows from the third condition of the
definition of simplicial maps, Definition~\ref{defnsimplMaps},
that $f = \phi$.
\end{proof}

The following theorem provides a necessary and sufficient condition to
create a simplicial approximation to a continuous function between two
simplicial complexes.

\begin{prop} \label{propExtSApprox}
Suppose that $K$ and $L$ are two simplicial complexes and that
$f:[K]\to [L]$ is a continuous function.  If $\phi$ is a map defined
on the set of vertices of $K$ with values in the set of vertices of
$L$, then $\phi$ can be extended to a simplicial approximation of
$f$ if and only if $f(\St(\VEC{x})) \subset \St(\phi(\VEC{x}))$ for
all vertices $\VEC{x}$ of $K$.
\end{prop}

\begin{proof}
Obviously, if $\phi$ can be extended to a simplicial approximation of
$f$, then we have from the definition of simplicial approximation that
$f(\St(\VEC{x})) \subset \St(\phi(\VEC{x}))$ for
all vertices $\VEC{x}$ of $K$.

Suppose that $f(\St(\VEC{x})) \subset \St(\phi(\VEC{x}))$ for
all vertices $\VEC{x}$ of $K$.  Consider a simplex
$(s) = (\VEC{x}_0,\VEC{x}_1,\ldots,\VEC{x}_m) \in K$.
Since $(s) \subset \St(\VEC{x}_j)$ for $0 \leq j \leq m$, we have
that $f\big( (s)\big) \subset f(\St(\VEC{x}_j)) \subset \St(\phi(\VEC{x}_j))$
for $0 \leq j \leq m$.  Thus $\displaystyle
\bigcap_{j=0}^m \St(\phi(\VEC{x}_j)) \neq \emptyset$.  Therefore,
there exists an open simplex $(t) \subset \St(\phi(\VEC{x}_j))$ for
$0 \leq j \leq m$.  We get that $\phi(\VEC{x}_j)$ for $0 \leq j \leq m$
are vertices of $[t]$.  Thus, we may define $\phi\big|_{[s]}$ by
$\displaystyle \phi\big|_{[s]}(\VEC{x}) = \sum_{j=0}^m a_j \phi(\VEC{x}_j)$ for
$\displaystyle \VEC{x} = \sum_{j=0}^k a_j \VEC{x}_j$ with $a_j >0$
and $\displaystyle \sum_{j=0}^k a_j = 1$.  In particular,
$\phi\big((s)\big) \subset [t]$.

Since $(s_1) = (s_2)$ or $(s_1) \cap (s_2) = \emptyset$ for all
simplices $(s_1)$ and $(s_2)$ of $K$, and $\phi\big|_{[s_1]}(\VEC{x})
= \phi\big|_{[s_2]}(\VEC{x})$ for $\VEC{x} \in [s_1]\cap [s_2]$,
we have that $\phi$ is a well defined simplicial map between $K$ and
$L$.
\end{proof}

\begin{theorem} \label{thESAf}
Suppose that $K$ and $L$ are two simplicial complexes and that
$f:[K]\to [L]$ is a continuous function.  Moreover, suppose that
$\{K_m\}_{m\in \NN}$ is a collection of subdivision of $K$ such that
$\mesh K_m \to 0$ as $m \to \infty$.  Then there exists $M>0$ such that
there exists a simplicial approximation $\phi_m:[K_n] \to [L]$ of $f$
for each $m \geq M$.
\end{theorem}

\begin{proof}
Since $\displaystyle \{\St(\VEC{y}) : \VEC{y} \ \text{is a vertex of} \ L\}$
is an open cover of $[L]$ with respect to the induced topology on
$[L]$ and $f:[K]\to [L]$ is continuous, we have that
$\displaystyle \{f^{-1}(\St(\VEC{y})) :
\VEC{y} \ \text{is a vertex of} \ L\}$ is an open cover of $[K]$ with
respect to the induce topology om $[K]$.  Since $[K]$ is compact,
there exists a number $\epsilon>0$, called the Lebesgue number, such
that any open ball in $[K]$ of radius less than or equal to $\epsilon$
is a subset $\displaystyle f^{-1}(\St(\VEC{y}))$ for some vertex
$\VEC{y}$ of $L$ \footnote{Since the number of vertices of $L$ is
finite, we could have proved the existence of $\epsilon$ without
making use of the theory of compact sets.}.

Choose $M$ such that $\mesh K_m < \epsilon/2$ for $m \geq M$.
Assume that $m \geq M$.  If ${s} \in K_m$, then
$\diam [s] \leq \mesh K_m < \epsilon/2$.  Hence, for each vertex
$\VEC{x}$ of $K_m$, we have that
$\displaystyle \St(\VEC{x}) \subset B_\epsilon(\VEC{x}) \subset 
f^{-1}(\St(\VEC{y}))$ for some vertex $\VEC{y}$ of $L$; namely,
\begin{equation} \label{scapproEq1}
  f(\St(\VEC{x})) \subset \St(\VEC{y})
\end{equation}
for some vertex $\VEC{y}$ of $L$.  If there are more then one vertex
$\VEC{y}$ of $L$ satisfying the previous statement, select one of
them.  Let $\phi_m(\VEC{x}) = \VEC{y}$.  This define a map from the set
of vertices of $K_m$ to the set of vertices of $L$.  We also have from
(\ref{scapproEq1}) that
$\displaystyle f(\St(\VEC{x})) \subset \St(\phi(\VEC{x}))$.
We can then use Proposition~\ref{propExtSApprox} to get a simplicial
approximation $\phi_m:[K_m] \to [L]$ of $f$.
\end{proof}

The previous theorem ensures that we can always find a simplicial
approximation as closed as we want to a continuous function between
two simplicial complexes.

\begin{cor}
Suppose that $K$ and $L$ are two simplicial complexes and that
$f:[K]\to [L]$ is a continuous function.  Given $\epsilon >0$, there
exist subdivisions $\tilde{K}$ and $\tilde{L}$ of $K$ and $L$
respectively, and a simplicial approximation
$\phi:[\tilde{K}]\to [\tilde{L}]$ of $f$, such that
$\|f - \phi\|_{[K]} < \epsilon$.
\end{cor}

\begin{proof}
As remarked after Proposition~\ref{propMeshMBS}, if
$\displaystyle L_j = L^{[j]}$ for $\displaystyle j\in \NNp$, then we
get a sequence of barycentric subdivisions of $L$ such that
$\mesh L_j \to 0$ as $j \to \infty$.  Thus there exists $j > 0$ such that
$\mesh L_j < \epsilon$.

Similarly, if $\displaystyle K_m = K^{[m]}$ for
$\displaystyle m\in \NNp$, then we get a sequence of barycentric
subdivisions of $M$ such that $\mesh K_m \to 0$ as
$m \to \infty$.  If we apply Theorem~\ref{thESAf} to
$f:[K] \to [L_j] = [L]$ using the collection of subdivisions
$\{K_m\}_{m\in \NNp}$ of $K$, then we find that there exists $m > 0$ and
a simplicial approximation $\phi:[K_m]\to [L_j]$ of $f$.
As remarked after Proposition~\ref{propNfmg}, we then have that
$\|f - \phi\|_{[K]} \leq \mesh L_j < \epsilon$.
We take $\tilde{K} = K_m$ and $\tilde{L} = L_j$.
\end{proof}

\subsection{Oriented Simplicial Homology} \label{subsOShomo}

From now on, when we refer to a simplicial simplex $K$, we assume that
a fixed indexing of the vertices of $K$ has been selected.  More
specifically, we assume that $\VEC{x}_0$, $\VEC{x}_1$, $\VEC{x}_2$,
\ldots are the vertices of $K$.

\begin{defn}
Let $[s]= [\VEC{x}_{j_0}, \VEC{x}_{j_1}, \ldots, \VEC{x}_{j_k}]$ be a
closed $k$-simplex.
An {\bfseries oriented $\mathbf{k}$-simplex}\index{Oriented $k$-simplex}
associated to $[s]$ is the pair $([s],\sigma)$ where
$\sigma: \{j_0,j_1,j_2,\ldots,j_k\} \to \{j_0,j_1,j_2,\ldots,j_k\}$
is a permutation.
\end{defn}

\begin{defn} \label{defnECordks}
Let $[s]= [\VEC{x}_{j_0}, \VEC{x}_{j_1}, \ldots, \VEC{x}_{j_k}]$
be a closed $k$-simplex.  We say that
two oriented $k$-simplices $([s],\sigma_1)$ and $([s],\sigma_2)$ are
{\bfseries equivalent}\index{Equivalent} if there exists an even
permutation $\beta:\{j_0,j_1,j_2,\ldots,j_k\} \to \{j_0,j_1,j_2,\ldots,j_k\}$
such that $\beta\circ \sigma_1 = \sigma_2$; namely,
$\beta(\sigma_1(j_i)) = \sigma_2(j_i)$ for $0 \leq i \leq k$.
\end{defn}

The notion of equivalence defined in the previous definition is an
equivalence relation that partitions the oriented $k$-simplices
associated to a closed $k$-simplex into two equivalence classes.

\begin{defn}
Let $[s] = [\VEC{x}_{j_0},\VEC{x}_{j_1}, \ldots, \VEC{x}_{j_k}]$
be a closed $k$-simplex.  We use the notations
$\os{\VEC{x}_{\sigma(j_0)}}{}{\VEC{x}_{\sigma(j_1)}}{}{\VEC{x}_{\sigma(j_k)}}$
or simply $\os{s_\sigma}{}{}{}{}$ to denote the equivalence class
associated to $([s],\sigma)$.
\end{defn}

If $[s]=[\VEC{x}_{j_0}, \VEC{x}_{j_1}, \ldots, \VEC{x}_{j_k}]$, then the
two equivalence classes of oriented $k$-simplices associate to $[s]$ are
$\os{\VEC{x}_{j_0}}{\VEC{x}_{j_1}}{\VEC{x}_{j_2}}{}{\VEC{x}_{j_k}}$ and
$\os{\VEC{x}_{j_1}}{\VEC{x}_{j_0}}{\VEC{x}_{j_2}}{}{\VEC{x}_{j_k}}$.
Note that
$\os{\VEC{x}_{j_0}}{\VEC{x}_{j_1}}{\VEC{x}_{j_2}}{}{\VEC{x}_{j_k}}$
is the equivalence class associated to $([s],\sigma_1)$ with
$\sigma_1 = \Id$ and
$\os{\VEC{x}_{j_1}}{\VEC{x}_{j_0}}{\VEC{x}_{j_2}}{}{\VEC{x}_{j_k}}$
is the equivalence class associated to $([s],\sigma_2)$ with
$\sigma_2$ is defined by
\[
\sigma_2(j_m) = \begin{cases}
j_1 & \quad \text{if} \ m = 0 \\
j_0 & \quad \text{if} \ m = 1 \\
j_m & \quad \text{otherwise}
\end{cases}
\]

\begin{defn}
Let $K$ be a simplicial complex and $R$ be an integral domain.
The {\bfseries module of $\mathbf{k}$-chains}\index{Module of $k$-Chains}
of $K$, denoted $C_k(K;R)$, is the free abelian group generated by the
equivalence classes of oriented $k$-simplices $\os{s_\sigma}{}{}{}{}$ for all
$(s) \in K$ where, given $(s) = (\VEC{x}_{j_0},\VEC{x}_{j_1},
\ldots, \VEC{x}_{j_k})$, we set
$\os{s_{\sigma_1}}{}{}{}{} = - \os{s_{\sigma_2}}{}{}{}{}$ if
$\beta\, \circ\, \sigma_1 = \sigma_2$ for an odd permutation
$\beta:\{j_0,j_1,j_2,\ldots,j_k\} \to \{j_0,j_1,j_2,\ldots,j_k\}$.
The elements of
$C_k(K;R)$ are called {\bfseries $\mathbf{k}$-chains}\index{$k$-Chain}
of $K$.
\end{defn}

For us, the integral domain $R$ will either be $\ZZ$ or $\RR$.

The elements $c \in C_k(K;R)$ are finite sums of the form
$\displaystyle c = \sum_{\substack{(s) \in K\\\dim(s)=k}} a_{(s)}
\os{s}{}{}{}{}{}$ with $a_{(s)} \in R$, or more explicitly
\[
  c = \sum_{(\VEC{x}_{j_0},\VEC{x}_{j_1},\ldots,\VEC{x}_{j_k}) \in K}
a_{(\VEC{x}_{j_0},\VEC{x}_{j_1},\ldots,\VEC{x}_{j_k})}
\os{\VEC{x}_{j_0}}{}{\VEC{x}_{j_1}}{}{\VEC{x}_{j_k}}
\]
for $a_{(\VEC{x}_{j_0},\VEC{x}_{j_1},\ldots,\VEC{x}_{j_k})}\in R$.

\begin{defn}
Let $K$ be a simplicial complex and $R$ be an integral domain.
If $(s) = (\VEC{x}_{j_0},\VEC{x}_{j_1},\ldots, \VEC{x}_{j_k}) \in K$
with $k>0$, then
the {\bfseries boundary}\index{Boundary} of $\os{s}{}{}{}{} \in C_k(K;R)$,
denoted $\partial_k \os{s}{}{}{}{}$, is defined by
\[
\partial_k \os{s}{}{}{}{} = \sum_{i=0}^k (-1)^i
\os{\VEC{x}_{j_0}}{}{\VEC{x}_{j_1}}{\widehat{\VEC{x}_{j_i}}}{\VEC{x}_{j_k}}
\in C_{k-1}(K;R) \ .
\]
(Figure~\ref{SimpHm1}).  If $(s) = (\VEC{x}_j) \in K$, then
$\partial_0 \os{\VEC{x}_j}{}{}{}{} = 0$.
\end{defn}

\pdfF{alg_top/simphm1}{Boundary of $k$-chains}{Examples of the
boundary of $k$-chains for the simplest $k$-chains.  For the $1$-chain
$\os{\VEC{x}_0}{}{}{}{\VEC{x}_1}$, we have that
$\partial_1 \os{\VEC{x}_0}{}{}{}{\VEC{x}_1} =
\os{\VEC{x}_0}{}{}{}{} - \os{\VEC{x}_1}{}{}{}{}$.
For the $2$-chain $\os{\VEC{x}_0}{\VEC{x}_1}{}{}{\VEC{x}_3}$, we have that
$\partial_2 \os{\VEC{x}_0}{\VEC{x}_1}{}{}{\VEC{x}_3} =
\os{\VEC{x}_1}{}{}{}{\VEC{x}_2} - \os{\VEC{x}_0}{}{}{}{\VEC{x}_2}
+ \os{\VEC{x}_0}{}{}{}{\VEC{x}_1} = 
\os{\VEC{x}_0}{}{}{}{\VEC{x}_1} + \os{\VEC{x}_1}{}{}{}{\VEC{x}_2}
+ \os{\VEC{x}_2}{}{}{}{\VEC{x}_0}$
as we have illustrated in the figure above.}{SimpHm1}

\begin{defn}
Let $K$ be a simplicial complex and $R$ an integral domain.
The {\bfseries boundary map}\index{Boundary Map}
$\partial_k: C_k(K;R) \to C_{k-1}(K;R)$ is define by
$\displaystyle \partial_k \bigg( \sum_{\substack{(s) \in K\\ \dim(s) = k}}
a_{(s)}\os{s}{}{}{}{} \bigg)
= \sum_{\substack{(s) \in K\\ \dim(s) =k}}  a_{(s)} \partial_k  \os{s}{}{}{}{}$
or more explicitly
\begin{align*}
&\partial_k \bigg( \sum_{(\VEC{x}_{j_0},\VEC{x}_{j_1},\ldots,\VEC{x}_{j_k}) \in K}
  a_{(\VEC{x}_{j_0},\VEC{x}_{j_1},\ldots,\VEC{x}_{j_k})}
  \os{\VEC{x}_{j_0}}{}{\VEC{x}_{j_1}}{}{\VEC{x}_{j_k}} \bigg) \\
&\hspace{7em} = \sum_{(\VEC{x}_{j_0},\VEC{x}_{j_1},\ldots,\VEC{x}_{j_k}) \in K}
  a_{(\VEC{x}_{j_0},\VEC{x}_{j_1},\ldots,\VEC{x}_{j_k})}
\partial_k  \os{\VEC{x}_{j_0}}{}{\VEC{x}_{j_1}}{}{\VEC{x}_{j_k}} \ .
\end{align*}
\end{defn}

A simple computation yields $\partial_{k-1} (\partial_k(c)= 0$ for all
$k$-chains $c$.  Note that $\partial_{k-1} (\partial_k(c)$ is the sum
of expressions of the form
\begin{align*}
&(-1)^{j_s+j_r}a_{\osscript{\VEC{x}_0}{}{\VEC{x}_1}{}{\VEC{x}_k}}
\os{\VEC{x}_{j_0}}{\VEC{x}_{j_1}}
{\ldots,\widehat{\VEC{x}_{j_r}}}{\widehat{\VEC{x}_{j_s}}}{\VEC{x}_{j_k}} \\
&\qquad\qquad + (-1)^{j_s+j_r-1}a_{\osscript{\VEC{x}_0}{}{\VEC{x}_1}{}{\VEC{x}_k}}
\os{\VEC{x}_{j_0}}{\VEC{x}_{j_1}}
{\ldots,\widehat{\VEC{x}_{j_r}}}{\widehat{\VEC{x}_{j_s}}}{\VEC{x}_{j_k}}
\end{align*}
that cancel each other.

\begin{defn}
Let $K$ be a simplicial complex and $R$ an integral domain,\\
$Z_k(K;R) = \{ c \in C_k(K;R) : \partial_k(c) = 0 \}$
and $B_k(K;R) = \{ \partial_{k+1}(c) : c \in C_{k+1}(K;R) \}$.
The {\bfseries (simplicial) $\displaystyle \mathbf{k^{th}}$ homology
module}\index{Homology Module!Simplicial $k^{th}$ Homology Module} of $K$ is the
free abelian quotient group $H_k(K;R) = Z_k(K;R) / B_k(K;R)$.  The
elements of $Z_k(K;R)$ are called {\bfseries cycles}\index{Cycle} and those of
$B_k(K;R)$ are called {\bfseries boundaries}\index{Boundary}.
The equivalence class of $H_k(K;R)$ associated to $c \in Z_k(K;R)$ is
denoted $[c]_K$.
\end{defn}

Suppose that $K$ and $L$ are two simplicial complexes and that
$f:[K] \to [L]$ is a homeomorphism that maps open simplices in $K$ to
open simplices in $L$.  The map $C(f): C_k(K;R) \to C_k(L;R)$ defined by
$\displaystyle C(f)\Big( \sum_{\substack{(s) \in K\\ \dim(s) =k}}
a_{(s)}\os{s}{}{}{}{} \Big)
= \sum_{\substack{(s) \in K\\ \dim(s) =k}} a_{(s)} \os{f((s))}{}{}{}{}$
is an isomorphism.
Since $C(f)$ commutes with $\partial_k$ for all $k$, we may use $C(f)$
to induce a map $H(f): H_k(K;R) \to H_k(L,R)$ because
$C(f)(Z_k(K;R)) \subset Z_k(L;R)$ and 
$C(f)(B_k(K;R)) \subset B_k(L;R)$.

\begin{prop} \label{propK0KZ}
Let $K$ be a simplicial complex and $R$ be an integral domain.  The
dimension of $H_0(K;R)$ is equal to the number of (connected) components
of $[K]$.  Namely, the number of independent generators of $H_0(K;R)$
is equal to the number of (connected) components of $[K]$.
\end{prop}

\begin{proof}[Proof (Sketch)]
Since $Z_0(K;R) = C_0(K;R)$ and all elements of $C_0(K;R)$
are of the form $\displaystyle c = \sum_{(\VEC{x}_j) \in K} a_{(\VEC{x}_j)}
\os{\VEC{x}_j}{}{}{}{}$ with $a_{(\VEC{x}_j)} \in R$,  it therefore
suffices to consider the linear independence of the equivalence
classes in $H_0(K;R)$ associated to
$\os{\VEC{x}_j}{}{}{}{} \in C_0(K;R)$ for $(\VEC{x}_j) \in K$ to
determine the dimension of $H_0(K;R)$.

If $(\VEC{x}_{j_0},\VEC{x}_{j_1}) \in K$, then
$\partial_1 \os{\VEC{x}_{j_0}}{}{}{}{\VEC{x}_{j_1}} = \os{\VEC{x}_{j_1}}{}{}{}{}
- \os{\VEC{x}_{j_0}}{}{}{}{}$.  Hence
$\os{\VEC{x}_{j_0}}{}{}{}{}$ and $\os{\VEC{x}_{j_1}}{}{}{}{}$ are
in the same equivalence classes of $H_0(K;R)$.
Inductively, we find that two equivalence classes of
$H_0(K;R)$, one associated to $\os{\VEC{x}_{j_0}}{}{}{}{}$ and one associated
$\os{\VEC{x}_{j_1}}{}{}{}{}$, are distinct if there is no sequence
$(x_{j_0},x_{k_1})$, $(x_{k_1},x_{k_2})$, $(x_{k_2},x_{k_3})$, \ldots,
$(x_{k_{m-1}},x_{k_m})$, $(x_{k_m},x_{j_1})$ of open $1$-simplices in
$K$.  This is only possible if $\VEC{x}_{j_0}$ and $\VEC{x}_{j_1}$
belongs to two distinct connected components of $[K]$.
\end{proof}

\begin{egg}
We consider the simplicial complex
\[
K = \{ (\VEC{x}_0), (\VEC{x}_1),
(\VEC{x}_2), (\VEC{x}_0,\VEC{x}_1), (\VEC{x}_1,\VEC{x}_2),
(\VEC{x}_0,\VEC{x}_2), (\VEC{x}_0,\VEC{x}_1,\VEC{x}_3) \} \ .
\]

\stage{a} It follows from Proposition~\ref{propK0KZ} that
$\displaystyle H_0(K^1;\ZZ) \cong \ZZ$.

\stage{b} We have
\begin{align*}
C_1(K^1;\ZZ) &= \left\{ a_{(\VEC{x}_0,\VEC{x}_1)}
\os{\VEC{x}_0}{}{}{}{\VEC{x}_1}
+ a_{(\VEC{x}_1,\VEC{x}_2)} \os{\VEC{x}_1}{}{}{}{\VEC{x}_2}
+ a_{(\VEC{x}_0,\VEC{x}_2)} \os{\VEC{x}_0}{}{}{}{\VEC{x}_2} : \right. \\
&\hspace{7em} \left. a_{(\VEC{x}_0,\VEC{x}_1)}, a_{(\VEC{x}_1,\VEC{x}_2)},
a_{(\VEC{x}_0,\VEC{x}_2)} \in \ZZ \right\}
\cong \ZZ^3
\end{align*}
and
\begin{align*}
&\partial_1 \left(a_{(\VEC{x}_0,\VEC{x}_1)} \os{\VEC{x}_0}{}{}{}{\VEC{x}_1}
+ a_{(\VEC{x}_1,\VEC{x}_2)} \os{\VEC{x}_1}{}{}{}{\VEC{x}_2}
+ a_{(\VEC{x}_0,\VEC{x}_2)} \os{\VEC{x}_0}{}{}{}{\VEC{x}_2} \right) \\
&\quad = a_{(\VEC{x}_0,\VEC{x}_1)}
\left(\os{\VEC{x}_1}{}{}{}{} - \os{\VEC{x}_0}{}{}{}{} \right)
+ a_{(\VEC{x}_1,\VEC{x}_2)}
\left(\os{\VEC{x}_2}{}{}{}{} - \os{\VEC{x}_1}{}{}{}{} \right)
+ a_{(\VEC{x}_0,\VEC{x}_2)}
\left(\os{\VEC{x}_2}{}{}{}{} - \os{\VEC{x}_0}{}{}{}{} \right) \\
&\quad = \left( a_{(\VEC{x}_0,\VEC{x}_1)} - a_{(\VEC{x}_1,\VEC{x}_2)} \right)
\os{\VEC{x}_1}{}{}{}{}
- \left( a_{(\VEC{x}_0,\VEC{x}_2)} + a_{(\VEC{x}_0,\VEC{x}_1)} \right)
\os{\VEC{x}_0}{}{}{}{}
+ \left(a_{(\VEC{x}_1,\VEC{x}_2)} + a_{(\VEC{x}_0,\VEC{x}_2)} \right)
\os{\VEC{x}_2}{}{}{}{} = 0
\end{align*}
if and only if
$a_{(\VEC{x}_0,\VEC{x}_1)} = a_{(\VEC{x}_1,\VEC{x}_2)} 
= - a_{(\VEC{x}_0,\VEC{x}_2)}$.  Hence
\[
Z_1(K^1;\ZZ)
= \left\{ a \left( \os{\VEC{x}_0}{}{}{}{\VEC{x}_1}
+ \os{\VEC{x}_1}{}{}{}{\VEC{x}_2}
- \os{\VEC{x}_0}{}{}{}{\VEC{x}_2} \right) : a \in \ZZ \right\}
\cong \ZZ \ .
\]
We also have that $\displaystyle B_1(K^1;\ZZ) = \{ 0 \}$ because
$\displaystyle C_2(K^1;\ZZ) = \{ 0 \}$.  Therefore
$\displaystyle H_1(K^1;\ZZ) = Z_1(K^1;\ZZ) / B_1(K^1;\ZZ) \cong \ZZ$.

\stage{c} It follows from Proposition~\ref{propK0KZ} that
$H_0(K;\ZZ) \cong \ZZ$. 

\stage{d} As before, we have
\[
Z_1(K;\ZZ)
= \left\{ a \left( \os{\VEC{x}_0}{}{}{}{\VEC{x}_1}
+ \os{\VEC{x}_1}{}{}{}{\VEC{x}_2}
- \os{\VEC{x}_0}{}{}{}{\VEC{x}_2} \right) : a \in \ZZ \right\}
\cong \ZZ \ .
\]
We have
\[
C_2(K;\ZZ) = \left\{ a \os{\VEC{x}_0}{\VEC{x}_2}{}{}{\VEC{x}_3}
: a \in \ZZ \right\} \cong \ZZ \ .
\]
Since
\begin{equation} \label{eggHkKZ2}
\partial_2 \left(a \os{\VEC{x}_0}{\VEC{x}_1}{}{}{\VEC{x}_2} \right)
= a \left(\os{\VEC{x}_1}{}{}{}{\VEC{x}_2} - \os{\VEC{x}_0}{}{}{}{\VEC{x}_2}
+ \os{\VEC{x}_0}{}{}{}{\VEC{x}_1} \right)
\end{equation}
for $a \in \ZZ$, we get
\[
B_1(K;\ZZ) = \left\{
a \left(\os{\VEC{x}_1}{}{}{}{\VEC{x}_2} - \os{\VEC{x}_0}{}{}{}{\VEC{x}_2}
+ \os{\VEC{x}_0}{}{}{}{\VEC{x}_1} \right) : a \in \ZZ \right\}
= Z_1(K^2;\ZZ) \cong \ZZ \ .
\]
Thus $H_1(K;\ZZ) = Z_1(K;\ZZ) / B_1(K;\ZZ) = 0$.

\stage{e} Finally, it follows from (\ref{eggHkKZ2}) that
$\partial_3 \left( a \os{\VEC{x}_1}{\VEC{x}_2}{}{}{\VEC{x}_3} \right) = 0$
if and only if $a = 0$.  So $Z_2(K;\ZZ) = \{ 0 \}$ and therefore
$H_2(K;\ZZ) = 0$.
\end{egg}

If $R = \RR$, then $C_k(K,\RR)$, $B_k(K,\RR)$ can be treated as
vector spaces over $\RR$.

We now define the Euler characteristic of a simplicial complex and
deduce some very interesting properties associated to
the Euler characteristic of a simplicial complex.

\begin{defn}   \label{defnBettiEulerC}
Let $K$ be a simplicial complex.
The {\bfseries $\displaystyle \mathbf{j^{th}}$ Betti
number}\index{$\displaystyle j^{th}$ Betti Number}
of $K$, denoted $\beta_j$, is the non-negative integer defined by
$\beta_j = \dim(H_j(K;\RR))$.
The {\bfseries Euler characteristic}\index{Euler Characteristic} of $K$,
denote $\Chi(K)$, is the integer defined by
$\displaystyle \Chi(K) = \sum_{j=0}^{\dim(K)} (-1)^j \beta_j$.
\end{defn}

Some of the readers may have recognized the famous Euler
characteristic formula introduced in undergraduate geometry courses
\cite{C}. 
Suppose that $K$ is the triangulation of the unit sphere
$\displaystyle S^2 \subset \RR^3$.   The Euler characteristic of this
triangulation $K$ of $\displaystyle S^2$ is given by $V - E + F$ where
$V$ is the number of vertices, $E$ is the number of edges, and $F$ is
the number of faces (i.e. triangles) of $K$.  It is a 
value independent of the triangulation of $\displaystyle S^2$.

\begin{theorem} \label{thmEulerAlpha}
Let $K$ be a simplicial complex and $\alpha_j$ be the number of
open $j$-simplices in $K$.  Then
$\displaystyle \Chi(K) = \sum_{j=0}^{\dim(K)} (-1)^j \alpha_j$.
\end{theorem}

\begin{proof}
Let $C_{-1}(K;\RR) = \{ 0 \}$.
Since $\partial_0(\os{\VEC{x}_j}{}{}{}{}) = 0$ for all vertices $\VEC{x}_j$
of $K$, we may also set $B_{-1}(K;\RR) = \{ 0 \}$.  Since
$\partial_i: C_i(K,\RR) \to C_{i-1}(K;\RR)$ is a linear operator for
$0 \leq i \leq \dim(K)$, we have
\[
\alpha_i = \dim(C_i(K;\RR)) = \dim(\KE(\partial_i)) + \dim(\IMG(\partial_i))
= \dim(Z_i(K;\RR)) + \dim(B_{i-1}(K;\RR))
\]
for $0\leq i \leq \dim(K)$.  We also have
\[
\beta_i = \dim(H_i(K;\RR)) = \dim\left(Z_i(K;\RR)/B_i(K;\RR)\right)
= \dim(Z_i(K;\RR)) - \dim(B_i(K;\RR))
\]
for $0\leq i \leq \dim(K)$.  Hence
\begin{align*}
\Chi(K) &= \sum_{i=0}^{\dim(K)} (-1)^i \beta_i
= \sum_{i=0}^{\dim(K)} (-1)^i \left( \dim(Z_i(K;\RR)) -
\dim(B_i(K;\RR)) \right) \\
&= \sum_{i=0}^{\dim(K)} (-1)^i \dim(Z_i(K;\RR))
+ \sum_{i=0}^{\dim(K)} (-1)^{i+1} \dim(B_i(K;\RR)) \\
&= \sum_{i=0}^{\dim(K)} (-1)^i \dim(Z_i(K;\RR))
+ \sum_{i=1}^{1+\dim(K)} (-1)^i \dim(B_{i-1}(K;\RR)) \\
&= \big( \dim(Z_0(K;\RR)) + \underbrace{\dim(B_{-1}(K;\RR))}_{=0} \big) \\
&\qquad
+ \sum_{i=1}^{\dim(K)} (-1)^i \left( \dim(Z_i(K;\RR)) + \dim(B_{i-1}(K;\RR))
\right) \\
&\qquad + (-1)^{1+\dim(K)} \underbrace{\dim(B_{\dim(K)}(K;\RR))}_{=0} \\
&= \sum_{i=0}^{\dim(K)} (-1)^i \left( \dim(Z_i(K;\RR)) + \dim(B_{i-1}(K;\RR))
\right)
= \sum_{i=0}^{\dim(K)} (-1)^i \alpha_i \ .
\end{align*}
Note that
$\dim(B_{\dim(K)}(K;\RR)) = 0$ because $C_{1+\dim(K)}(K;\RR) = \{ 0 \}$.
\end{proof}

\subsection{Reduced and Relative Homology} \label{ssectSimplRDh}

Let $K$ be a simplicial complex and $R$ be an integral domain.
We have by definition that $\partial_0(c) = 0$ for all 
$c \in C_0(K;R)$.  Namely, we are assuming that the
image of $\partial_0$ is in the space $C_{-1}(K;R) = \{ 0 \}$.

We replace $C_{-1}(K;R) = \{0\}$ and the boundary operator
$\partial_0 : C_0(K;R) \to C_{-1}(K;R)$ 
to obtain a slightly new collection of homology modules.
Let
\[
C_k^\sharp(K;R) = \begin{cases}
R & \quad \text{if} \ k = -1 \\
C_k(K;R) & \quad \text{if} \ k \neq -1
\end{cases}
\]
Also, let $\displaystyle \partial_k^\sharp = \partial_k$ for $k \neq 0$ and
\[
\partial_0^\sharp \Big( \sum_{(\VEC{x}_j)\in K}
a_{(\VEC{x}_j)}\os{\VEC{x}_j}{}{}{}{} \Big)
= \sum_{(\VEC{x}_j) \in K} a_{(\VEC{x}_j)}
\]
for $a_{(\VEC{x}_j)} \in R$.
It is not difficult to prove that
$\displaystyle \partial_0^\sharp \circ \partial_1^\sharp= 0$.
We can then proceed as we did before to define
$\displaystyle Z_k^\sharp(X;R)$, $\displaystyle B_k^\sharp(X;R)$
and $\displaystyle H_k^\sharp(X;R)$.

\begin{defn}
Let $K$ be a simplicial complex and $R$ be an integral domain.
The quotient $\displaystyle H_k^\sharp(K;R) =
Z_k^\sharp(X;R) / B_k^\sharp(X;R)$ is called the
{\bfseries reduced (simplicial) $\displaystyle \mathbf{k^{th}}$ homology
module}\index{Homology Module!Reduced (Simplicial) $k^{th}$ Homology Module}
of $X$.
\end{defn}

Again, let $K$ be a simplicial complex and $R$ be an integral domain.
Suppose that $L$ is a simplicial subcomplex of $K$.  Recall that this
means that $L \subset K$ and $L$ is itself a simplicial complex.
Assuming that we keep the same indexing of the vertices of $L$ as the
one used for $K$, then $C_k(L;R)$ is a submodule of $C_k(K;R)$.

\begin{defn}
Let $L$ be a simplicial subcomplex of a simplicial complex $K$ and $R$
be an integral domain.
The {\bfseries module of relative chains}\index{Module of Relative Chains}
of $K \mod L$, denoted $C_k(K,L;R)$, is the free abelian group defined as
$\displaystyle C_k(K,L;R) = C_k(K;R)/C_k(L;R)$.
\end{defn}

Since $\partial_k:C_k(L;R) \to C_{k-1}(L;R)$, we may define 
the operator $\overline{\partial}_k:C_k(K,L;R) \to C_{k-1}(K,L;R)$
as $\overline{\partial}_k(\relC[K,L]{c}) = \relC[K,L]{\partial_k c}$ for all
$\relC[K,L]{c} \in C_k(K,L;R)$ where $\relC[K,L]{c}$ denotes the equivalence
class in $C_k(K,L;R)$ associated to $c \in C_k(K;R)$.
We get the following commutative diagram.
\[
\xymatrix@C+2ex{
C_k(K;R) \ar[d]^{\partial_k} \ar[r]^-{\pi_k} & C_k(K,L;R)
\ar[d]^{\overline{\partial}_k} \\
C_{k-1}(K;R) \ar[r]^-{\pi_{k-1}} & C_{k-1}(K,L;R)
}
\]
where $\pi_k:C_k(K;R) \to C_k(K,L;R)$ is the projection defined
by $\pi_k(c) = \relC[K,L]{c}$ for $c \in C_k(K;R)$.

\begin{defn}
Let $L$ be a simplicial subcomplex of a simplicial complex $K$ and $R$
be an integral domain.
The {\bfseries relative (simplicial) $\displaystyle \mathbf{k^{th}}$ homology
module}\index{Homology Module!Relative (Simplicial) $k^{th}$ Homology Module} of
$K \mod L$ is the free abelian group
$\displaystyle H_k(K,L;R) = \KE(\overline{\partial}_k)
/\IMG(\overline{\partial}_{k-1})$.
The equivalence class in $H_k(K,L;R)$ associated to
$\relC[K,L]{c} \in Z_k(K,L;R) = \KE(\overline{\partial}_k)$ is denoted
$[c]_{K,L}$.
\end{defn}

It is possible to develop the theory of reduced and relative homology
as it is done in \cite{MUat} but we will only do so in the context of
singular homology later on.
Since singular homology seems to be more accessible and
user friendly than simplicial homology (though the latest is more
visual), we will only study reduced and relative homology
in the context of singular homology in Sections~\ref{sectSingHom}.

\section{Simplicial Cohomology} \label{sectSimplCohom}

We only provide a very brief introduction to simplicial cohomology.
We give just enough information to reach our goal of providing a
relation between the de Rham cohomology on a manifold $S$ and the
simplicial cohomology on a simplicial complex $K$ associated (in some
way to be defined) to $S$.

\subsection{Cohomology Modules}

\begin{defn}
Let $K$ be a simplicial complex.  For $0 \leq k \leq \dim(K)$, let
$\displaystyle C^k(K;\RR)$ denote the dual space of $C_k(K;\RR)$.
Namely, $\displaystyle C^k(K;\RR)$ is the vector space of all linear
functionals on $C_k(K;\RR)$.  The elements of $\displaystyle C^k(K;\RR)$
are called {\bfseries cochaines}\index{Cochaines}. 
\end{defn}

\begin{defn}
Let $K$ be a simplicial complex.  The adjoint of the boundary map
$\partial_k: C_k(K;\RR) \to C_{k-1}(K;\RR)$ is the map
$\displaystyle \dfC_{k-1}: C^{k-1}(K;\RR) \to C^k(K;\RR)$ defined
by
\begin{equation} \label{simplCoOpEq1}
\big(\dfC_{k-1}(\phi)\big)(c) = \phi\big(\partial_k(c)\big)
\end{equation}
for all $c \in C_k(K;\RR)$ and $\displaystyle \phi \in C^{k-1}(K;\RR)$.
The map $\displaystyle \dfC_k$ is called
the {\bfseries coboundary operator}\index{Coboundary Operator}.
\end{defn}

It is clear that $\displaystyle \dfC_k$ is well and uniquely
defined by (\ref{simplCoOpEq1}).  It is easy to prove using
$\partial_{k+1} \circ \partial_{k+2} = 0$ that
$\displaystyle \dfC_{k+1} \circ \dfC_k = 0$.

\begin{defn}
Let $K$ be a simplicial complex,
$\displaystyle Z^k(K;\RR) = \{ \phi \in C^k(K;\RR) :
\dfC_k \phi = 0 \}$
and $\displaystyle B^k(K;\RR) = \{ \dfC_{k-1} \phi : \phi \in
C^{k-1}(K;\RR) \}$.
The {\bfseries $\displaystyle \mathbf{k^{th}}$ cohomology
module}\index{Cohomology Module!$k^{th}$ Cohomology Module}
of $K$ is the
quotient $\displaystyle H^k(K;\RR) = Z^k(K;\RR) / B^k(K;\RR)$.
The elements of $\displaystyle Z^k(K;\RR)$ are called
{\bfseries cocycles}\index{Cocycle} and those of
$\displaystyle B^k(K;\RR)$ are called
{\bfseries coboundaries}\index{Coboundary}.
The equivalence class in $\displaystyle H^k(K;R)$ associated to
$\displaystyle \phi \in Z^k(K;R)$ is denoted $[\phi]_K$.
\end{defn}

\begin{theorem} \label{thmHupeHdk}
Let $K$ be a simplicial complex.  Then $\displaystyle H^k(K;\RR)$ is
isomorphic to $\displaystyle (H_k(K;\RR))^\ast$.
\end{theorem}

\begin{proof}
The map $\displaystyle h:H^k(K;\RR) \to (H_k(K;\RR))^\ast$ defined by
$h([\phi]_K) = \phi$ for $\displaystyle \phi \in Z^k(K;\RR)$ is well
defined because
\begin{align*}
\phi \in Z^k(K;\RR) & \Rightarrow \dfC_k(\phi) = 0
\Rightarrow \phi(\partial_{k+1}(c)) = (\dfC_k \phi)(c) =  0
\ \text{for all} \ c \in C_{k+1}(K;\RR) \\
&\Rightarrow \phi = 0 \ \text{on} \ B_k(K;\RR)
\Rightarrow \phi \in (H_k(K;\RR))^\ast \ .
\end{align*}

We have that $h$ is one-to-one.  Suppose that $[\phi_1]_K = [\phi_2]_K$.
Then $\displaystyle \phi_1 - \phi_2 = \psi \in B^k(K;\RR)$.  Hence
$h([\phi_1]_K) = h([\phi_2]_K) + h([\psi]_K) = h([\phi_2]_K)$
because
\begin{align*}
\psi \in B^k(K;\RR)
&\Rightarrow \psi = \dfC_{k-1} \eta \ \text{for some}
\ \eta \in C^{k-1}(K;\RR) \\
&\Rightarrow \psi(c) = (\dfC_{k-1} \eta)(c)
= \eta(\partial_k(c)) = 0 \ \text{for all} \ c \in Z_k(K;\RR) \\
&\Rightarrow \psi = 0 \ \text{on}\ Z_k(K;\RR)
\Rightarrow h([\psi]_K) = 0 \ .
\end{align*}

We have that $h$ is onto.  Given $\displaystyle \phi \in (H_k(K;\RR))^\ast$,
then $\displaystyle [\phi]_K \in H^k(K;\RR)$ because
\begin{align*}
\phi(c) = 0 \ \text{for all}\ c \in B_k(K;\RR)
&\Rightarrow (\dfC_k \phi)(c) = \phi(\partial_{k+1}(c)) = 0
\ \text{for all} \ c \in C_{k+1}(K;\RR) \ \text{since} \\
&\qquad \ \partial_{k+1}(c) \in B_k(K;\RR) \ \text{for all}
\ c \in C_{k+1}(K;\RR) \\
&\Rightarrow \dfC_k \phi = 0 \Rightarrow \phi \in Z^k(K;\RR) \ .
\end{align*}
Thus $h([\phi]_K) = \phi$.

We leave it to the reader to proof that $h$ is a linear mapping.

Hence, the map $h$ is an isomorphism between $\displaystyle H^k(K;\RR)$ and
$\displaystyle (H_k(K;\RR))^\ast$.
\end{proof}

It is possible to describe explicitly the action of the coboundary
operator $\displaystyle \dfC_k$.  Suppose that $K$ is a
simplicial complex.   Given an open $k$-simplex $(s)$ in $K$, we defined
$\displaystyle \phi_{\osscript{s}{}{}{}{}} \in C^k(K;\RR)$ by
\begin{equation} \label{dfnBCkKR}
\phi_{\osscript{s}{}{}{}{}}(\os{t}{}{}{}{}) =
\begin{cases}
1 & \quad \text{if} \ \os{t}{}{}{}{} = \os{s}{}{}{}{} \\    
-1 & \quad \text{if} \ \os{t}{}{}{}{} = -\os{s}{}{}{}{} \\    
0 & \quad \text{otherwise}
\end{cases}
\end{equation}
for all open $k$-simplices $(t)$ in $K$.
If $\big\{ \os{s_1}{}{}{}{}, \os{s_2}{}{}{}{}, \ldots, \os{s_m}{}{}{}{} \big\}$
is a basis of $C_k(K;\RR)$, then
$\big\{ \phi_{\osscript{s_1}{}{}{}{}}, \phi_{\osscript{s_2}{}{}{}{}}, \ldots,
\phi_{\osscript{s_m}{}{}{}{}} \big\}$ is a basis of $\displaystyle C^k(K;\RR)$.

\begin{prop}
We have for all open $k$-simplices
$(s) = (\VEC{x}_0,\VEC{x}_1,\ldots,\VEC{x}_k) \in K$ that
\begin{equation} \label{pakpoEq1}
\dfC_k (\phi_{\osscript{\VEC{x}_0}{}{\VEC{x}_1}{}{\VEC{x}_k}})
= \sum_{\substack{(\VEC{y}) \in K\\
(\VEC{y},\VEC{x}_0,\VEC{x}_1,\ldots, \VEC{x}_k) \in K}}
\phi_{\osscript{\VEC{y}}{\VEC{x}_0}{\VEC{x}_1}{}{\VEC{x}_k}} \ .
\end{equation}
\end{prop}

\begin{proof}
Since $C_{k+1}(K;\RR)$ is generated by oriented $(k+1)$-simplices, we
only have to prove (\ref{pakpoEq1}) for an arbitrary ordered
$(k+1)$-simplex $\os{\VEC{z}_0}{}{\VEC{z}_1}{}{\VEC{z}_{k+1}}$.

By definition,
\begin{align*}
\big(\dfC_k (\phi_{\osscript{\VEC{x}_0}{}{\VEC{x}_1}{}{\VEC{x}_k}})
\big)(\os{\VEC{z}_0}{}{\VEC{z}_1}{}{\VEC{z}_{k+1}})
&= \phi_{\osscript{\VEC{x}_0}{}{\VEC{x}_1}{}{\VEC{x}_k}}
\big(\partial_{k+1}(\os{\VEC{z}_0}{}{\VEC{z}_1}{}{\VEC{z}_{k+1}})\big) \\
&= \phi_{\osscript{\VEC{x}_0}{}{\VEC{x}_1}{}{\VEC{x}_k}}
\bigg(\sum_{j=0}^{k+1} (-1)^j
\os{\VEC{z}_0}{}{\VEC{z}_1}{\widehat{\VEC{z}_j}}{\VEC{z}_{k+1}}\bigg) \\
&= \sum_{j=0}^{k+1} (-1)^j
\phi_{\osscript{\VEC{x}_0}{}{\VEC{x}_1}{}{\VEC{x}_k}}
\big(\os{\VEC{z}_0}{}{\VEC{z}_1}{\widehat{\VEC{z}_j}}{\VEC{z}_{k+1}}\big) \ .
\end{align*}
However
\begin{align*}
&\phi_{\osscript{\VEC{x}_0}{}{\VEC{x}_1}{}{\VEC{x}_k}}
\big(\os{\VEC{z}_0}{}{\VEC{z}_1}{\widehat{\VEC{z}_j}}{\VEC{z}_{k+1}}\big) \\
& \qquad = \begin{cases}
1 & \quad \text{if}
\ \os{\VEC{z}_0}{}{\VEC{z}_1}{\widehat{\VEC{z}_j}}{\VEC{z}_{k+1}}
= \os{\VEC{x}_0}{}{\VEC{x}_1}{}{\VEC{x}_k} \\
-1 & \quad \text{if}
\ \os{\VEC{z}_0}{}{\VEC{z}_1}{\widehat{\VEC{z}_j}}{\VEC{z}_{k+1}}
= - \os{\VEC{x}_0}{}{\VEC{x}_1}{}{\VEC{x}_k} \\
0 & \quad \text{otherwise}
\end{cases} \\
& \qquad = \begin{cases}
1 & \quad \text{if}
\ \os{\VEC{z}_0}{}{\VEC{z}_1}{}{\VEC{z}_{k+1}}
= (-1)^j \os{\VEC{z}_j}{\VEC{x}_0}{\VEC{x}_1}{}{\VEC{x}_k} \\
-1 & \quad \text{if}
\ \os{\VEC{z}_0}{}{\VEC{z}_1}{}{\VEC{z}_{k+1}}
= (-1)^{j+1} \os{\VEC{z}_j}{\VEC{x}_0}{\VEC{x}_1}{}{\VEC{x}_k} \\
0 & \quad \text{otherwise}
\end{cases} \\
& \qquad = \begin{cases}
(-1)^j \phi_{\osscript{\VEC{z}_j}{\VEC{x}_0}{\VEC{x}_1}{}{\VEC{x}_k}}
\big(\os{\VEC{z}_0}{}{\VEC{z}_1}{}{\VEC{z}_{k+1}}\big)
& \quad \text{if}
\ \os{\VEC{z}_j}{\VEC{x}_0}{\VEC{x}_1}{}{\VEC{x}_k} \\
& \quad = \pm \os{\VEC{z}_0}{}{\VEC{z}_1}{}{\VEC{z}_{k+1}} \\
0 & \quad \text{otherwise}
\end{cases}
\end{align*}
Thus (\ref{pakpoEq1}) is true for 
$\os{\VEC{z}_0}{}{\VEC{z}_1}{}{\VEC{z}_{k+1}}$.
\end{proof}

\subsection{Relative Cohomology}

It is also natural to define relative $k$-cohomology modules.  This is
another topic that we only briefly introduce but that could be greatly
developed.  We will not do so in the context of simplicial theory but
will treat it a little bit more seriously in
Subsection~\ref{ssectRCsingT} in the context of singular theory.

\begin{defn}
Let $L$ be a simplicial subcomplex of a simplicial complex $K$, and $R$
be an integral domain.  The adjoint of the boundary operator 
$\overline{\partial}_k : C_k(K,L;R) \to C_{k-1}(K,L;R)$ is the operator
$\displaystyle \overline{\dfC}_{k-1} : \big(C_{k-1}(K,L;R)\big)^\ast
\to \big(C_k(K,L;R)\big)^\ast$ defined
\begin{equation} \label{relsimplCoOpEq1}
\big(\overline{\dfC}_{k-1}(\phi)\big)([c])
= \phi\big(\overline{\partial}_k([c])\big)
\end{equation}
for all $\relC[K,L]{c} \in C_k(K,L;R)$ and
$\displaystyle \phi \in \big(C_{k-1}(K,L;R)\big)^\ast$.
\end{defn}

It is clear that $\displaystyle \overline{\dfC}_k$ is well
and uniquely defined by (\ref{relsimplCoOpEq1}).

\begin{defn}
Let $L$ be a simplicial subcomplex of a simplicial complex $K$, and $R$
be an integral domain.
The {\bfseries relative $\displaystyle \mathbf{k^{th}}$ cohomology
module}\index{Cohomology Module!Relative $k^{th}$ Cohomology Module} of
$K \mod L$ is the
quotient space $\displaystyle H^k(K,L;R) = \KE(\overline{\dfC}_k)/
\IMG(\overline{\dfC}_{k-1})$.
\end{defn}

\subsection{Cup Product}

There is a feature of cohomology theory that is equivalent to the
wedge product of differential forms on manifolds.  It is called the
cup product.

\begin{defn}
Let $K$ be a simplicial complex and $R$ be an integral domain.
The {\bfseries cup product}\index{Cup Product} of
$\displaystyle \phi_1 \in C^{k_1}(K;R)$ with
$\displaystyle \phi_2 \in C^{k_2}(K;R)$ is the linear functional
$\displaystyle \phi_1 \cup \phi_2 \in C^{k_1+k_2}(K;R)$ defined first by
$(\phi_1 \cup \phi_2)(\os{\VEC{x}_0}{}{\VEC{x}_1}{}{\VEC{x}_{k_1+k_2}})
= \phi_1(\os{\VEC{x}_0}{}{\VEC{x}_1}{}{\VEC{x}_{k_1}})
\,\phi_2(\os{\VEC{x}_{k_1}}{}{\VEC{x}_{K_1+1}}{}{\VEC{x}_{k_1+k_2}})$
for all $(\VEC{x}_0,\VEC{x}_1, \ldots,\VEC{x}_{k_1+k_2}) \in K$ and
after extended linearly to $\displaystyle C^{k_1+k_2}(K;R)$.
\end{defn}

It is not hard to prove that the cup product is a bilinear map, that
it is associative and that the identity element is the linear functional
$\displaystyle \delta \in C^0(K;R)$ defined by
$\delta(\os{\VEC{x}_i}{}{}{}{}) = 1$ for all vertices $\VEC{x}_i$ of $K$.
We define the cup product of
$\displaystyle [\phi_1]_K \in H^{k_1}(K;R)$ and
$\displaystyle [\phi_2]_K \in H^{k_2}(K;R)$ as
$\displaystyle [\phi_1]_K\cup[\phi_2]_K = [\phi_1 \cup \phi_2]_K$.

We will not justify these definitions and deduce their properties.  We
will prove later that simplicial homology and simplicial cohomology on
the simplex $K$ are respectively isomorphic to singular homology and
singular cohomology on $[K]$.  It is in the context of singular
theory that we will prove that the cup product is well defined, and
present other results about the cup product. There is effectively a
lot more that can be said about the cup product.

\subsection{Relation Between Simplicial Cohomology and
de Rham Cohomology}  \label{ssectSCandCM}

\begin{defn}  \label{defnSTriangleM}
A triple $(S,K,h)$ is a {\bfseries smoothly triangulated
manifold}\index{Smoothly Triangulated Manifold} if
\begin{enumerate}
\item $S$ is a manifold of class $\displaystyle C^\infty$,
\item $K$ is a simplicial complex,
\item $h:[K] \to S$ is a homeomorphism,
\item there exist, for every
$(s) = (\VEC{x}_{j_0},\VEC{x}_{j_1}, \ldots, \VEC{x}_{j_p}) \in K$, an
open subset $U_{[s]}$ of the $k$-dimensional affine space
$\displaystyle A_{[s]} =
\{\VEC{x}_{j_0} + \sum_{i=1}^p y_i (\VEC{x}_{j_i}-\VEC{x}_{j_0}) :
y_i \in \RR \ \text{for} \ 1 \leq i \leq p \}$
and a map $h_{[s]}:U_{[s]} \to S$ of class $\displaystyle C^\infty$
such that $[s] \subset U_{[s]}$ and $h_{[s]}(\VEC{x}) = h(\VEC{x})$
for all $\VEC{x} \in [s]$ (Figure~\ref{SimpHm2}), and
\item $h_{[s]}(U_{[s]})$ is a submanifold of $S$ of class
$\displaystyle C^\infty$.
\end{enumerate}
\end{defn}

\pdfF{alg_top/simphm2}{Smoothly triangulated manifolds}{We consider
the simplicial complex\\ $K = \{(\VEC{x}_0),(\VEC{x}_1),(\VEC{x}_2),
(\VEC{x}_0,\VEC{x}_1),(\VEC{x}_0,\VEC{x}_2),(\VEC{x}_1,\VEC{x}_2)\}$
and the circle $\displaystyle S^1$ inscribed in $[K]$.  The maps $h$
is represented by the black arrows.  A possible open set
$U_{[\VEC{x}_0,\VEC{x}_1]}$ and map $h_{[\VEC{x}_0,\VEC{x}_1]}$ is
provided with $h_{[\VEC{x}_0,\VEC{x}_1]}$ on
$U_{[\VEC{x}_0,\VEC{x}_1]} \setminus [\VEC{x}_0,\VEC{x}_1]$ 
represented by the blue arrows.}{SimpHm2}

\begin{defn} \label{defnjthBC}
Let $K$ be a simplicial complex and $\VEC{x}_j$ for $0 \leq j \leq q$
be the vertices of $K$.  The
{\bfseries $\displaystyle \mathbf{j^{th}}$ barycentric
coordinates}\index{Barycentric Coordinates!$j^{th}$ Barycentric Coordinates} of
$\VEC{x} \in [K]$, denoted $b_j(\VEC{x})$, is defined by
\[
b_j(\VEC{x}) = \begin{cases}
0 & \hspace{1.5em} \text{if} \ x \not\in \St(\VEC{x}_j) \\
\begin{array}{l}
\text{The $j^{th}$ barycentric} \\[-0.3em]
\text{coordinate of} \ \VEC{x} \in (s)
\end{array} & \quad
\text{if}\ \VEC{x} \in (s) \subset \St(\VEC{x}_j)
\end{cases}
\]
\end{defn}

\begin{rmk}
The $\displaystyle j^{th}$ barycentric coordinates given in
Definition~\ref{defnjthBC} have several interesting properties.
\begin{enumerate}
\item  We have that $b_j:[K]\to \RR$ is continuous.  For instance, suppose
that $K$ is the simplicial complex consisting of all the
faces of $[\VEC{x}_2,\VEC{x}_3,\VEC{x}_5]$ and the open simplices
$(\VEC{x}_0)$, $(\VEC{x}_1)$, $(\VEC{x}_0,\VEC{x}_1)$ and
$(\VEC{x}_1,\VEC{x}_2)$ as represented in the following figure.
\pdfbox{alg_top/simphm3}
$\St(\VEC{x}_2)$ is represented by the continuous lines and the
shaded area.  Given\\
$\VEC{x} \in (\VEC{x}_2,\VEC{x}_3,\VEC{x}_4)$, we have that
$\displaystyle \VEC{x} = \sum_{j=2}^4 a_j \VEC{x}_j$
with $\displaystyle \sum_{j=2}^4 a_j = 1$.
The function $b_2$ is continuous with respect to the $a_j$ in the open set
$(\VEC{x}_2,\VEC{x}_3,\VEC{x}_4)$.  Therefore, if
$\VEC{x}$ converges to $\tilde{\VEC{x}} \in [\VEC{x}_3,\VEC{x}_4]$, then
$a_3$ and $a_4$ converge to $\tilde{a}_3$ and $\tilde{a}_4$
respectively where
$\displaystyle \tilde{\VEC{x}} = \sum_{j=3}^4 \tilde{a}_j \VEC{x}_j$
with $\displaystyle \sum_{j=3}^4 \tilde{a}_j = 1$.
Thus $a_3 + a_4 \to 1$ and so $b_2(\VEC{x}) = a_2$ converges to $0$.
Hence $b_2$ is continuous along $[\VEC{x}_3,\VEC{x}_4]$.  Likewise,
if $\VEC{x}$ converges to $\breve{\VEC{x}} \in [\VEC{x}_2,\VEC{x}_3]$, then
$a_2$ and $a_3$ converge to $\breve{a}_2$ and $\breve{a}_3$
respectively where
$\displaystyle \breve{\VEC{x}} = \sum_{j=2}^3 \breve{a}_j \VEC{x}_j$
with $\displaystyle \sum_{j=2}^3 \breve{a}_j = 1$.
Thus $b_2(\VEC{x}) = a_2$ converges to $\breve{a}_2 = b_2(\breve{\VEC{x}})$.
Hence $b_2$ is continuous along $[\VEC{x}_2,\VEC{x}_3]$.  Proceeding
in a similar way, we can prove that $b_2$ is continuous on all of
$[K]$.  Likewise, $b_1$ and $b_3$ are continuous on all of
$[K]$.
\item For all $\VEC{x} \in [K]$, we have that
$\displaystyle \VEC{x} = \sum_{j=0}^q b_j(\VEC{x}) \VEC{x}_j$ with
$b_j(\VEC{x}) \geq 0$ and $\displaystyle \sum_{j=0}^q b_j(\VEC{x}) = 1$.
More precisely, since
$\VEC{x} \in (\VEC{x}_{j_0},\VEC{x}_{j_1}, \ldots, \VEC{x}_{j_p}) \in K$
for some $j_i \in \{0,1,2,\ldots,q\}$ and $0 \leq i \leq p$, we get
that $\displaystyle \VEC{x} = \sum_{i=0}^p b_{j_i}(\VEC{x}) \VEC{x}_{j_i}$
with $b_{j_i}(\VEC{x}) > 0$ and
$\displaystyle \sum_{i=0}^p b_{j_i}(\VEC{x}) = 1$.  Moreover,
$b_j(\VEC{x}) = 0$ if $j \not\in \{ j_0,j_1,\ldots,j_p\}$.
\end{enumerate}
\end{rmk}

\begin{defn} \label{defnStarOS}
Let $K$ be a simplicial complex and $(s)$ be an open simplex in $K$.
The {\bfseries star}\index{Star} of $(s)$, denoted $\St\big((s)\big)$, is
defined by
$\displaystyle \St\big((s)\big)
= \bigcup_{\substack{(t)\, \in K \text{\ such that}
\\(s) \text{ is a face of } [t]}} (t)$.
\end{defn}

Recall that, by definition of a simplicial complex, if $(s) \in K$,
then all the faces of $[s]$ are also in $K$; in particular,
$(s) \subset \St\big((s)\big)$ because $(s)$ is a face of $[s]$.

\begin{rmk}
Below are some of the properties of the star of an open simplex as
given in Definition~\ref{defnStarOS}.  Some these properties are easy
to prove and proving them is left to the reader.
\begin{enumerate}
\item For a $0$-simplex $(\VEC{x}) \in K$,
we have that $\St\big((\VEC{x})\big) = \St(\VEC{x})$, the star of the
vertex $\VEC{x}$ of $K$ as given in Definition~\ref{defnStarV}.
\item If $(s)$ is an open simplex of $K$, then $\St\big((s)\big)$ is
an open set in $[K]$.  This follows from the fact that $\St\big((s)\big)$
is the union of open simplices of $[K]$.
\item If $(s) = (\VEC{x}_{j_0},\VEC{x}_{j_1},\ldots,\VEC{x}_{j_p}) \in K$
where the $\VEC{x}_{j_i}$ for $0 \leq i \leq p$ are vertices of $K$, then
$\VEC{x} \in \St\big((s)\big)$ if and only if
$b_{j_i}(\VEC{x}) \neq 0$ for $0\leq i \leq p$.
To verify this claim, it suffices to note that if
$\VEC{x} \in \St\big((s)\big)$ then $\VEC{x} \in (t)$ such that $(s)$
is a face of $[t]$.  Thus
$(t) = (\VEC{x}_{j_0},\VEC{x}_{j_1}, \ldots, \VEC{x}_{j_q})$ with
$q\geq p$ and $j_i \not\in \{j_0,j_1, \ldots, j_p\}$ for $p < i \leq q$.
Hence $\displaystyle \VEC{x} = \sum_{i=0}^q b_{j_i}(\VEC{x}) \VEC{x}_{j_1}$
with $b_{j_i}(\VEC{x}) > 0$ for $0 \leq i \leq q$
and $\displaystyle \sum_{i=0}^q b_{j_i}(\VEC{x}) = 1$.  
\item It follows from (3) that
$\displaystyle [K]\setminus \St\big((s)\big) =
\big\{ \VEC{x} \in K : b_{j_i}(\VEC{x}) = 0 \ \text{for some} \ i\in
\{0,1,\ldots,p\} \big\}$.
\item If $(s_1)$ and $(s_2)$ are two open $q$-simplices of a simplicial
complex $K$, and $(s_1) \neq (s_2)$, then
$[s_2] \subset [K]\setminus \St\big((s_1)\big)$.  This follows from
the fact that all faces of a closed $q$-simplex, other than the open
$q$-simplex itself, are open $j$-simplices with $j<q$.  So $(s_1)$
cannot be a face of $[s_2]$.
\end{enumerate}
\end{rmk}

Suppose that $(S,K,h)$ is a smoothly triangulated manifold.  We would
like to find a relation between the $\displaystyle k^{th}$ cohomology
module of $S$ defined in Definition~\ref{defnqdeRhamCM} and the
$\displaystyle k^{th}$ cohomology group $\displaystyle H^k(K;\RR)$ of
$K$.  For that, we will define a map
$\displaystyle F_k : C^\infty\big(S,\Omega^k(S)\big) \to C^k(K,\RR)$
such that the following diagram commutes.
\[
\xymatrix{
C^\infty\big(S,\Omega^k(S)\big) \ar[r]^-{\df{}}\ar[d]_{F_k} &
C^\infty\big(S,\Omega^{k+1}(S)\big) \ar[d]^{F_{k+1}} \\
C^k(K;\RR) \ar[r]_-{\dfC_k} & C^{k+1}(K;\RR)
}
\]
Suppose that $\omega$ is a differential $k$-form on $S$.  To define
$\displaystyle F_k(\omega) \in C^k(K;\RR)$, it is enough to define
$F_k(\omega)$ on the equivalence classes of oriented
$k$-simplices of $K$ and use the linearity of $F_k(\omega)$ to extend
its definition to all of $\displaystyle C_k(K;\RR)$.
Given $\os{s}{}{}{}{} \in C_k(K;\RR)$ with
$(s) = (\VEC{x}_{j_0},\VEC{x}_{j_1},\ldots,\VEC{x}_{j_k}) \in K$, there
exists an open subset $U_{[s]}$ and a smooth mapping $h_{[s]}:U_{[s]} \to S$
that satisfy Definition~\ref{defnSTriangleM}.  We set
\[
  F_k(\omega)(\os{s}{}{}{}{}) = \int_{[s]} h_{[s]}^\ast(\omega) \ .
\]
We may consider that $[s]$ is a subset of the $k$-dimensional manifold
$U_{[s]}$.  This manifold can be described by only one local chart;
namely, $(W_{[s]},U_{[s]},\phi_{[s]})$ where
\begin{align*}
\phi_{[s]}:\RR^k &\to \{\VEC{x}_{j_0}
+ \sum_{i=1}^k y_i (\VEC{x}_{j_i}-\VEC{x}_{j_0}) :
y_i \in \RR \ \text{for} \ 1 \leq i \leq k \} \\
\VEC{y} & \mapsto \VEC{x}_{j_0} + \sum_{i=1}^k y_i (\VEC{x}_{j_i}-\VEC{x}_{j_0})
\end{align*}
and $\displaystyle W_{[s]} = \phi_{[s]}^{-1}(U_{[s]})$.
Hence, the local representation of $\displaystyle h_{[s]}^\ast(\omega)$
is given by $\displaystyle \phi_{[s]}^\ast(h_{[s]}^\ast(\omega)) =
g_{[s]} \df{y_1}\wedge \df{y_2} \wedge \ldots \wedge \df{y_q}$ where
$g_{[s]}:W_{[s]} \to \RR$ is of class $\displaystyle C^\infty$ and
\[
F_k(\omega)(\os{s}{}{}{}{}) = \int_{W_{[s]}} g_{[s]}
\df{y_1}\wedge \df{y_2} \wedge \ldots \wedge \df{y_k} \ .
\]

\begin{prop}
We have that $\displaystyle \dfC_k \circ F_k = F_{k+1} \circ \df{}$.
\end{prop}

\begin{proof}
It suffices to use Stokes' theorem.  Given a differential $k$-form on
$S$ and $\os{s}{}{}{}{} \in C_k(K,\RR)$, we have
\begin{align*}
F_{k+1}(\df{\omega})(\os{s}{}{}{}{})
&= \int_{[s]} h_{[s]}^\ast(\df{\omega})
= \int_{[s]} \df{\left(h_{[s]}^\ast(\omega)\right)}
= \int_{\partial [s]} h_{[s]}^\ast(\omega) \\
&= F_k(\omega)(\partial_k \os{s}{}{}{}{})
= \dfC_k(F_k(\omega))(\os{s}{}{}{}{}) \ .
\end{align*}
Stokes' theorem, Theorem~\ref{stokesStokes}, was used for the third
equality.  Moreover, $\partial [s]$ must be interpreted according to
Stokes' theorem. Hence the orientation on $\partial [s]$ is the
induced orientation from the orientation on $[s]$.  It is in fact the
orientation given by $\partial_k \os{s}{}{}{}{}$.
\end{proof}

The next theorem is the main result of this section.

\begin{theorem}[De Rham's Theorem] \label{DeRhamThm}
Let $(S,K,h)$ be a smoothly triangulated manifold and
$\displaystyle \tilde{F}_k:H^k(S) \to H^k(K;\RR)$ be the map defined by
$\displaystyle \tilde{F}_k([\omega])([c]) = [F_k(\omega)(c)]_K$ for the
closed differential $k$-form $\omega$ on $S$ and $k$-chains $c$ in $Z_k(K;\RR)$.
Then $\tilde{F}_k$ is an isomorphism for $0 \leq k \leq \dim(S)$.
\end{theorem}

The proof of this theorem is very long but it is not hard. We only
sketch the proof of this famous theorem.  The reader can find the
detailed proof in \cite{ST}.

\begin{proof}[Proof (Sketch)]
\stage{Claim 1} There exist linear maps
$\displaystyle G_k: C^k(K;\RR) \to C^\infty\big(S,\Omega^k(S)\big)$
for $0 \leq k \leq \dim(S)$ such that:
\begin{enumerate}
\item $\displaystyle \df{}\circ G_k = G_{k+1}\circ \dfC_k$ for
$0 \leq k < \dim(S)$.
\item $F_k \circ G_k = \Id$.
\item If $\displaystyle \phi \in C^0(K,\RR)$ satisfies
$\phi(\os{\VEC{x}}{}{}{}{}) = 1$
for all $0$-simplices $(\VEC{x}) \in K$, then $G_0(\phi)$ is the
differential $0$-form (i.e.\ a function on $S$) defined by
$G_0(\phi)(\VEC{x}) = 1$ for all $\VEC{x} \in S$.
\item If $\os{s}{}{}{}{}$ is an oriented $k$-simplex of $K$, then
$G_k(\phi_{\osscript{s}{}{}{}{}})$ is null in an open set containing $[K]
\setminus \St\big((s)\big)$, where $\phi_{\osscript{s}{}{}{}{}}$ is
defined in (\ref{dfnBCkKR}).
\end{enumerate}

\stage{i} Suppose that $K$ is a simplicial complex of dimension $q$
and that $\VEC{x}_0$, $\VEC{x}_1$, \ldots, $\VEC{x}_p$ are the
vertices of $K$.  We define a partition of unity subordinated to the open
cover $\big\{ \St(\VEC{x}_j) \big\}_{0\leq j \leq p}$
of $[K]$.

Let
$A_j = \big\{ \VEC{x} \in [K] : b_j(\VEC{x}) \geq 1/(q+1) \big\}$
and
$B_j = \big\{ \VEC{x} \in [K] : b_j(\VEC{x}) \leq 1/(q+2) \big\}$
for $0 \leq j \leq p$, where $b_j$ is the barycentric coordinate
function defined in Definition~\ref{defnjthBC}.

Since $A_j$ is a closed subset of the compact set $[K]$, we have
that $A_j$ is compact for $0 \leq j \leq p$.

We have that $[K]\setminus \St(\VEC{x}_j) \subset B_j$
and $A_j \subset C_j = [K] \setminus B_j \subset \St(\VEC{x}_j)$
for $0 \leq j \leq p$ as can be seen in the figure below.
\pdfbox{alg_top/derham1}
In the figure above, the dotted area represents $\St(\VEC{x}_j)$.
The collection $\{ A_j\}_{0\leq j \leq p}$ is a closed cover of
$[K]$.  Given $\VEC{x} \in [K]$, we have that $\VEC{x}$ is an element
of an open $m$-simplex
$(s) = (\VEC{x}_{j_1},\VEC{x}_{j_2}, \ldots, \VEC{x}_{j_m})$
of $K$ with $m \leq q$.  By definition of the barycentric coordinates,
we have that $b_{j_i}(\VEC{x}) > 0$ for $0\leq i \leq m$ and
$\displaystyle \sum_{i=0}^m b_{j_i}(\VEC{x}) = 1$.  Therefore, we must
have that $b_{j_i}(\VEC{x}) \geq 1/(m+1) \geq 1/(q+1)$ for some
$0 \leq i \leq m$.  Thus $\VEC{x} \in A_{j_i}$.

Given $0 \leq j \leq p$, there exists according to
Proposition~\ref{topolProp2} an open set
$\displaystyle V_j \subset \RR^n$ such that
$\displaystyle  A_j \subset V_j \subset \overline{V}_j
\subset \RR^n \setminus B_j$.
Using partition of unity for instance, we can select a function
$\displaystyle f_j \in C^\infty(\RR^n)$ such that
$0 \leq f_j(\VEC{x}) \leq 1$ for all $\displaystyle \VEC{x} \in \RR^n$,
$f_j(\VEC{x}) = 1$ for all $\VEC{x} \in A_j$ and $f_j(\VEC{x}) = 0$
for all $\displaystyle \VEC{x} \not\in V_j$.
It follows from the previous paragraph that, given any $\VEC{x} \in [K]$,
there exists $0 \leq j \leq p$ such that $f_j(\VEC{x}) = 1$.
Let $\displaystyle g = \sum_{j=0}^pf_j$ on $\displaystyle \RR^n$.  We
have that $g(\VEC{x}) \geq 1$ for all $\VEC{x} \in [K]$.  Thus
$g(\VEC{x}) \geq 1/2 > 0$ for all $\VEC{x}$ in an open set $W \supset [K]$. 
Let $g_j = f_j/g$  on $W$ for $0 \leq j \leq p$.  Hence
$\displaystyle g_j\in C^\infty(W)$ and
$\displaystyle \sum_{j=0}^pg_j = 1$ on $[K]$.  Moreover,
$\supp g_j\Big|_{[K]} \subset \supp f_j\Big|_{[K]}
\subset C_j \subset \St(\VEC{x}_j)$ for $0 \leq j \leq p$.

The collection $\{g_j\}_{0\leq j \leq p}$ is the partition of unity
subordinated to the open cover\\
$\big\{ \St(\VEC{x}_j) \big\}_{0\leq j \leq p}$
of $[K]$ that we are looking for.

\stage{ii} Since $G_k$ is linear, we only have to define $G_k$ for
$\phi_{\osscript{s}{}{}{}{}}$ where $\os{s}{}{}{}{}$ is an
oriented $k$-simplices of $K$.  Suppose that
$\os{s}{}{}{}{}= \os{\VEC{x}_{j_0}}{}{\VEC{x}_{j_1}}{}{\VEC{x}_{j_k}}$.
We set
\[
G_k(\phi_{\osscript{s}{}{}{}{}})
= k! \sum_{i=0}^k (-i)^i g_{j_i} \df{g_{j_0}} \wedge \df{g_{j_1}}
\wedge \ldots \wedge \widehat{\df{g_{j_i}}} \wedge \ldots \wedge \df{g_{j_k}}
\ .
\]
Note that $G_k(\phi_{\osscript{s}{}{}{}{}})$ defines
a differential $k$-form on $W \supset [K]$.  We can use the maps
$h_{[s]]}$ provided by the smoothly triangulated manifold $(S,K,h)$ to
transport $G_k(\phi_{\osscript{s}{}{}{}{}})$ to $S$ \footnote{In
\cite{ST}, they simplify the writing of the proof by assuming that
$[K] = S$.  In other words, they project $K$ on $S$ using the maps $h_s$.}. 

The proof given in \cite{ST} that $G_k$ satisfies all four conditions
listed above requires, among several results, Stokes' theorem on chains.

\stage{Claim 2}
If $\omega$ is a closed differential $k$-form on $S$ such that
$\displaystyle F_k(\omega) = \dfC_k(\phi)$ for some
$\displaystyle \phi \in C^{k-1}(K;R)$,
then there exists a differential $(k-1)$-form $\rho$ on $S$ such that
$F_{k-1}(\rho) = \phi$ and $\df{\rho} = \omega$.

The proof of this claim is in two parts.

\stage{First Part}
Given an open $q$-simplex $(s) \in K$, let
$\displaystyle B_{(s)}
= \bigcup_{\substack{(t) \text{ an open face of }[s]\\\text{with } \dim (t) < q}}
(t)$.

We prove the following statements.
\begin{enumerate}
\item Assuming $q>0$, if $\omega$ is a closed differential $k$-form defined
in an open neighbourhood of $B_{(s)}$ in $\displaystyle \RR^n$
where $(s) \in K$ is an open $q$-simplex. then there exists a closed
differential $k$-form $\rho$ defined in an open neighbourhood of $[s]$
such that $\rho = \omega$ in the open neighbourhood of $B_{(s)}$.  If
$q=k+1$, then we must also require that
$\displaystyle \int_{\partial [s]} \omega =0$.
\item Assuming $q,k>0$, if $\omega$ is a closed differential $k$-form
defined in an open neighbourhood of $[s]$ in $\displaystyle \RR^n$ where
$(s) \in K$ is an open $q$-simplex, and $\rho$ is a differential $(k-1)$-form
defined in an open neighbourhood of $B_{(s)}$ and such
that $\df{\rho} = \omega$ in that neighbourhood of $B_{(s)}$, then
there exists a differential $(k-1)$-form $\nu$ defined in an open
neighbourhood of $[s]$ such that $\nu = \rho$ in an open neighbourhood
of $B_{(s)}$ and $\df{\nu} = \omega$ in a neighbourhood of $[s]$.
If $q=k$, then we must also
require that $\displaystyle \int_{\partial [s]} \rho = \int_{[s]} \omega$.
\end{enumerate}

Note that the two extra conditions in the two previous statement are
required because we get from Stokes' theorem that
\[
\int_{\partial [s]} \omega = \int_{\partial [s]} \rho
= \int_{[s]} \df{\rho} = \int_{[s]} 0 = 0
\]
if $\rho$ exists in the first statement and $q = k+1$, and
\[
\int_{[s]} \omega = \int_{[s]} \df{\nu}
= \int_{\partial [s]} \nu = \int_{\partial [s]} \rho
\]
if $\nu$ exists in the second statement and $q=k$.

The proof of the two statements above is an iterative proof.  We first
prove that (1) is true for $k=0$.  Next, we prove that if (1) is true
for $k=i$, then (2) is true for $k=i+1$.  Finally, we prove that
if (2) is true for $k=i+1$, then (1) is true for $k=i+1$.  Hence,
iterating from $i=0$ up to $i = \dim(S)$ shows that (1) and (2) are
true for all values of $k$

To prove (1) for $k=0$, we note that $\omega$ is a function on an open
neighbourhood of $B_{(s)}$ such that $\displaystyle \pdydx{\omega}{x_i} = 0$ in
this open neighbourhood for all $i$.  Thus $\omega$ is constant on the
components of $B_{(s)}$.  If $q>1$, then $B_{(s)}$ is
connected and so $\omega(\VEC{x}) = c$, a constant, for all
$\VEC{x} \in B_{(s)}$.  Therefore, we may set $\rho = c$ in an opne
neighbourhood of $[s]$.  If $q=1$, then
$[s] = [\VEC{x}_{j_1},\VEC{x}_{j_2}]$ for some vertices 
$\VEC{x}_{j_1}$ and $\VEC{x}_{j_2}$ of $K$, and
$B_{(s)} = \{\VEC{x}_{j_1},\VEC{x}_{j_2}\}$.  But since $q = k+1$, we
have that
$\displaystyle 0 = \int_{\partial [s]} \omega = \omega(\VEC{x}_{j_1}) - 
\omega(\VEC{x}_{j_0})$.  Thus $\omega(\VEC{x}_{j_1}) =
\omega(\VEC{x}_{j_0}) = c$, a constant.  Therefore, we may again set
$\rho =c$ in an open neighbourhood of $[s]$.

To prove that (1) for $k=i$ implies (2) for $k=i+1$, we may assume that
$\omega$ is a closed differential $k$-form in a connected open set
$U \supset [s]$ because $[s]$ is connected.  It then follows from 
Corollary~\ref{corHqg0e0} that $\omega = \df{\mu_1}$ for some
differential $(k-1)$-form $\mu_1$ on $U$ \footnote{We are using the fact that
$\displaystyle H^k(U) = H^k(\RR^n) = 0$ for $k>0$ because $U$ is
diffeomorphic to $\displaystyle \RR^n$.}.  Since $\mu_1$ may not be
equal to $\rho$ in a neighbourhood of $B_{(s)}$, we consider
$\mu_2 = \mu_1 - \rho$.  We have that
$\df{\mu_2} = \df{\mu_1} - \df{\rho} = \omega - \omega = 0$ in a
neighbourhood of $B_{(s)}$ and, if $q = k$, we have from Stokes'
theorem that
\[
\int_{\partial [s]} \mu_2
= \int_{\partial [s]} \mu_1 - \int_{\partial [s]} \rho
= \int_{[s]} \df{\mu_1} - \int_{\partial [s]} \rho
= \int_{[s]} \omega - \int_{\partial [s]} \rho
= 0
\]
by assumption.  Therefore, we may use (1) with $\mu_2$ to get a closed
differential $k$-form $\mu_3$ defined in an open neighbourhood of
$[s]$ such that $\mu_3 = \mu_2$ in a neighbourhood of $B_{(s)}$.
Then $\nu = \mu_1 - \mu_3$ because $\nu = \rho$ and
$\df{\nu} = \df{\mu_1} - \df{\mu_3} = \omega$ in a neighbourhood of 
$B_{(s)}$.

To prove that (2) for $k=i+1$ implies (1) for $k=i+1$ is a little bit
trickier.  Suppose that
$(s) = (\VEC{x}_{j_0},\VEC{x}_{j_1},\ldots,\VEC{x}_{j_q})$.  Let
$(t) = = (\VEC{x}_{j_1},\ldots,\VEC{x}_{j_q})$ and
$F_{(s)} = B_{(s)}\setminus (t)$.  Since $F_{(s)}$ is
star-shaped, we may assume that the open neighbourhood of $F_{(s)}$
where $\omega$ is defined is also star-shaped.  Hence, we may use
Poincar Lemma to get a differential $(k-1)$-form $\mu$ defined in
this neighbourhood such that $\df{\mu} = \omega$.

If $q>1$, then we can show that (2) can be used with $\rho$ replaced by
$\mu$ and $(s)$ by $(t)$ to get a differential $(k-1)$-form $\nu$
defined in an open neighbourhood of $[t]$ such that
$\nu = \mu$ in an open neighbourhood of $B_{(t)}$ and $\df{\nu} = \omega$
in a neighbourhood of $[t]$.  Consider the differential $k$-form
defined by
\[
\tau = \begin{cases}
\mu & \quad \text{on the neighbourhood of}\ F_{(s)} \\
\nu & \quad \text{on the neighbourhood of}\ [t]
\end{cases}
\]
where the neighbourhoods may have to be shrunk a little if necessary.
We have that $\df{\tau} = \omega$ in a neighbourhood of $B_{(s)}$.
If $q=1$, then we also have that there is a differential $k-1$-form $\tau$
defined in $B_{(s)}$ such that $\df{\tau} = \omega$.  In this case, we
use the fact that the neighbourhood of $B_{(s)}$ may be chosen to be
the union of two disconnect open sets; one about each of the two
vertices in $B_{(s)}$. 

To complete the proof, we select a real valued function
$\displaystyle f \in C^\infty(\RR^n)$ such that $f=1$ in an open
neighbourhood of $B_{(s)}$ and $f=0$ outside the open neighbourhood
where $\tau$ is defined.  Then $\rho = \df{(f \tau)}$ satisfies (1)
because it is obviously exact and 
\[
\df{\rho} = \df{f} \wedge \tau + f \wedge \df{\tau} = \df{\tau} = \omega
\]
in the open neighbourhood where $f =1$.

\stage{Second Part}
The second part of the proof of Claim 2 used the second statement
of the previous part to construct a sequence of differential $(k-1)$-forms
$\rho_0$, $\rho_1$, $\rho_2$, \ldots , $\rho_{\dim(S)}$ such that
$\rho_j$ is defined in an open neighbourhood of the $j$-skeleton
$\displaystyle [K^j]$ of $K$,
$\df{\rho_j} = \omega$ in this open neighbourhood of $\displaystyle [K^j]$,
$\rho_j = \rho_{j-1}$ in an open neighbourhood of $\displaystyle [K^{j-1}]$
for $0 < j \leq \dim(S)$, and $F_j(\rho_j) = \phi$ if $j = k-1$.
We have that $\rho$ in Claim 2 is given by $\rho = \rho_{\dim(S)}$.

The construction of the $\rho_j$ is done by induction on the order $j$
of the skeletons of $K$.

For $j=0$, the skeleton $\displaystyle K^0$ is a
set of discrete points.  We may assume that the open neighbourhood of
$\displaystyle [K^0]$ is the union of disjoint open balls $B_j$ with
$\VEC{x}_j$ being the only vertex of $K$ in $B_j$.  Using Poincar
Lemma on each open ball, we get a differential $(k-1)$-form
$\tilde{\rho}_0$ on the open neighbourhood of $\displaystyle [K^0]$ such that
$\df{\tilde{\rho}_0} = \omega$.
If $k-1 > 0$, then we may take $\rho_0 = \tilde{\rho}_0$.
If $k-1 = 0$, then we have to make sure that
$F_0(\rho_0) = \phi$ on $\displaystyle [K^0]$.
For each vertex $\displaystyle \VEC{x}_j$ of $K$, set $\displaystyle
a_j = \phi\big( (\VEC{x}_j)\big) - \int_{(\VEC{x}_j)}  \tilde{\rho}_0
= \phi\big( (\VEC{x}_j)\big) - \tilde{\rho}_0(\VEC{x}_j)$.
Then $\rho_0$ defined in the open neighbourhood of
$\displaystyle [K^0]$ by $\rho_0\big|_{B_j} = \tilde{\rho}_0 + a_j$ for all $j$
satisfies all the required conditions.  In particular,
$\displaystyle F_0(\rho_0)\big((\VEC{x}_j)\big) = \int_{(\VEC{x}_j)}
\big( \tilde{\rho}_0 + a_j \big) = \phi\big( (\VEC{x}_j)\big)$ for all
$\VEC{x}_j$.

We assume that we have constructed $\rho_{j-1}$ according to the
requirements.

We have that $\omega$ is a closed differential $k$-form defined in
an open neighbourhood of a closed $j$-simplex $[s]$ and $\rho_{j-1}$ is a
differential $(j-1)$-form defined in an open neighbourhood of
$B_{(s)}$ such that $\df{\rho_{j-1}} = \omega$ in this open
neighbourhood of $B_{(s)}$.  If $k = j$, then
\[
\int_{[s]} \omega
= F_k(\omega)\big(\os{s}{}{}{}{}\big)
= \dfC_k(\phi)\big(\os{s}{}{}{}{}\big)
= \phi\big( \partial_k \os{s}{}{}{}{} \big)
= F_{j-1}(\rho_{j-1})\big( \partial_k \os{s}{}{}{}{}\big)
= \int_{\partial_k [s]} \rho_{j-1}
\]
where the second equality comes from the assumption in Claim 2 and the
fourth equality from our hypothesis of induction.
Hence the second statement of the first part can be used to get a
differential $(k-1)$-form $\tilde{\rho}_{(s)}$ in an open neighbourhood
$U_{(s)}$ of $[s]$ such that $\tilde{\rho}_{[s]} = \rho_{j-1}$ in an open
neighbourhood of $B_{(s)}$ and $\df{\tilde{\rho}_{(s)}} = \omega$ in
$U_{(s)}$.   We define $\tilde{\rho}_j$ in a neighbourhood
of $\displaystyle [K^j]$ by
$\tilde{\rho}_j\big|_{U_{(s)}} = \tilde{\rho}_{(s)}$ for
all open $j$-simplices $(s)$ in $K$.   The differential $(k-1)$-form
$\tilde{\rho}_j$ is well defined by because
$\tilde{\rho}_{(s_1)} = \tilde{\rho}_{(s_2)} = \rho_{j-1}$ in an open
neighbourhood of the intersection of $[s_1]$ and $[s_2]$ for all
open $j$-simplices $(s_1)$ and $(s_2)$ in $K$.

If $k-1 \neq j$, then we may define $\rho_j$ in a neighbourhood
of $\displaystyle [K^j]$ as $\rho_j = \tilde{\rho}_j$.

If $k-1 = j$, then we use a trick similar to the one that we have used in
the case $j=0$ to ensure that $F_j(\rho_j) = \phi$ on $\displaystyle C^j(K;R)$.
Let $\displaystyle \alpha_j = \phi - F_{k-1}(\tilde{\rho}_{k-1}) \in C^j(K;R)$.
We define $\rho_j$ in a neighbourhood $U$ of $\displaystyle [K^j]$ as
$\rho_j = \tilde{\rho}_j\big|_U + G_{k-1}(\alpha_j)\big|_U$.

We have from Claim 1 that $G_q(\phi_{\osscript{t}{}{}{}{}}) = 0$
in an open neighbourhood of $K \setminus \St\big( (t) \big)$ for all
open $q$-simplices $(t)$ in $K$.  Thus
$G_q(\phi_{\osscript{t}{}{}{}{}}) = 0$ in a neighbourhood of
$\displaystyle [K^{q-1}]$ for all open $q$-simplices $(t)$ in $K$.
Since all elements of $\displaystyle C^q(K;R)$ are linear combinations
of elements of the form $\phi_{\osscript{t}{}{}{}{}}$ for $(t)$ an
open $q$-simplex in $K$, we get that $G_q(\phi) = 0$ in a
neighbourhood of $\displaystyle [K^{q-1}]$ for all
$\displaystyle \phi \in C^q(K;R)$.  Thus, for $q=k-1 = j$, we
get that
\[
\df{\rho_j} = \df{\tilde{\rho}_j} + \df{\big( G_{k-1}(\alpha_j)\big)}
= \df{\tilde{\rho}_j} + G_k\big(\dfC_{k-1}(\alpha_j)\big)
= \df{\tilde{\rho}_j} = \omega
\]
in a neighbourhood of $\displaystyle [K^{j}]$, and
\[
\rho_j = \tilde{\rho}_j + G_{k-1}(\alpha_j)
= \tilde{\rho}_j = \rho_{j-1}
\]
in a neighbourhood of $\displaystyle [K^{j-1}]$.  Moreover,
\[
F_j(\rho_j)(\os{s}{}{}{}{})
= F_j(\tilde{\rho}_j)(\os{s}{}{}{}{})
+ F_j\big( G_{k-1}(\alpha_j)(\os{s}{}{}{}{}) \big)
= \phi(\os{s}{}{}{}{}) - \alpha_j(\os{s}{}{}{}{})
+ \alpha_j(\os{s}{}{}{}{})
= \phi(\os{s}{}{}{}{})
\]
for all $j$-simplices $(\os{s}{}{}{}{})$, where we have use (2) of
Claim 1 to get the second equality.

\stage{Conclusion}
It follows from (1) of Claim 1 that
$\displaystyle G_k(Z^k(K;\RR)) \subset Z^k(S)$ where
$\displaystyle Z^k(S)$ is the set of closed differential $k$-forms, and
$\displaystyle G_k(B^k(K;\RR)) \subset B^k(S)$ where
$\displaystyle B^k(S)$ is the set of exact differential $k$-forms.
Thus $G_k$ defines a map, that we denote $\tilde{G}_k$, from
$\displaystyle H^k(K;\RR)$ to $\displaystyle H^k(S)$.  Hence, it
follows from (2) that $\tilde{G}_k$ is a right inverse for
$\tilde{F}_k$ and so $\tilde{F}_k$ is onto.

It follows from Claim 2 that, if $\displaystyle \omega \in Z^k(S)$ and
$\displaystyle F_k(\omega) \in B^k(K;\RR)$, then
$\displaystyle \omega \in B^k(S)$.  Since $F_k$ is a linear map, this
proves that $\tilde{F}_k([\omega]) = [0]_K$ implies that $[\omega] = [0]$.
Thus $F_k$ is one-to-one.  This completes the proof that $\tilde{F}_k$ is an
isomorphism.
\end{proof}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "notes"
%%% End:
