\begin{thebibliography}{ABC}

\bibitem{A} Ralph Abraham, Jerrold E. Marsden and Tudor Ratiu,
{\bfseries Manifolds, Tensor Analysis, and Applications},
Addison-Wesley Publishing Company, 1983.

\bibitem{B} Bourbaki, N. {\bfseries Variétés différentielles et
analytiques}, Springer-Verlag, 2007.

\bibitem{CBV} Ruel V. Churchill, James W. Brown and Roger F. Verhey,
{\bfseries Complex Variables and Application}, McGraw-Hill, Inc., 1974.

\bibitem{C} H. S. M. Coxeter,  {\bfseries Introduction to Geometry,
2$^{nd}$ Edition}, John Wiley \& Sons, Inc., 1969.

\bibitem{Du} James Dugundji, {\bfseries Topology}, Allyn and Bacon, Inc.,
1978.

\bibitem{Fl} Wendell Fleming, {\bfseries Functions of Several Variables},
UTM, Springer-Verlag, 1977.

\bibitem{F} Gerald B. Folland, {\bfseries Advanced Calculus}, Prentice Hall,
2002.

\bibitem{G} Marvin J. Greenberg, {\bfseries Euclidean and non-Euclidean
Geometries: Development and History}, W. H. Freeman and Company, 1980.

\bibitem{GH} Marvin J. Greenberg and John R. Harper, {\bfseries
Algebraic Topology: A First Course (revised)}, Addison-Wesley Publishing
Company, 1981.

\bibitem{GP} Victor Guillemin and Alan Pollack, {\bfseries Differential
geometry}, Prentice Hall, 1974.

\bibitem{He} I. N. Herstein, {\bfseries Topics in Algebra}, John Wiley
\& Sons, Inc. 1975.

\bibitem{Hu} Thomas W. Hungerford, {\bfseries Algebra}, Holt, Rinehart
and Winston, Inc., 1974.

\bibitem{La} Serge Lang, {\bfseries Linear Algebra, 2$^{nd}$ Edition},
Addison-Wesley Publishing Company, 1971.

\bibitem{LaDM} Serge Lang, {\bfseries Differential Manifolds},
Addison-Wesley Publishing Company, 1972.

\bibitem{LJM} John M. Lee, {\bfseries Riemannian Manifolds: An
Introduction to Curvature}, Springer-Verlag, 1997.

\bibitem{LWAJ} W. A. J.Luxemburg, {\bfseries ArselÀ's Dominated
Convergence Theoem}, American Mathematics Monthly, Vol.\ 78,
970-979, New York, 1971.

\bibitem{MH} J. E. Marsden and M. J. Hoffman, {\bfseries Elementary Classical
Analysis, 2$^{nd}$ Edition}, W. H. Freeman and Company, 1993.

\bibitem{MT} J. E. Marsden and A. J. Tromba, {\bfseries Vector Calculus,
2$^{nd}$ Edition}, W. H. Freeman and Company, 1981.

\bibitem{Mi} J. Milnor, {\bfseries Topology from a Differential Viewpoint},
University of Virginia Press, 1965.

\bibitem{MU} J. R. Munkres, {\bfseries Topology, A First Course},
Printice-Hall, Inc., 1975.

\bibitem{MUat} J. R. Munkres, {\bfseries Elements of Algebraic Topology},
Addison-Wesley Publishing Company, 1984.

\bibitem{Ra} Tibor Radö. {\bfseries Length and Area}, American
Mathematical Society, Volume XXX, New York, 1948.

\bibitem{R} Walter Rudin, {\bfseries Principle of Mathematical Analysis,
3$^{rd}$ Edition}, McGraw-Hill, 1978.

\bibitem{Rrca} Walter Rudin, {\bfseries Real and Complex Analysis,
2$^{md}$ Edition}, McGraw-Hill, 1974.

\bibitem{ST} I. M. Singer and J. A. Thorpe, {\bfseries Lecture Notes
on Elementary Topology and Geometry}, UTM, Springer-Verlag, 1967.

\bibitem{S} Micheal Spivak, {\bfseries Calculus on Manifolds},
Benjamin/Cummings Publishing Comp. 1965.

\bibitem{Sv1} Micheal Spivak, {\bfseries A Comprehensive Introduction
to Differential Geometry, Volume One}, Publish or Perish, 2005.

\end{thebibliography}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "notes"
%%% End: 
