\nonumchapter{Preface}

The first part of these lecture notes is a complement to Spivak's
{\bfseries Calculus on Manifolds} \cite{S}.   We review some of the
topics in the book, adding some additional information and results.
We also reproduce (in some cases with major differences) some of the
proofs that were sketchy.  Finally, a large number of problems, many
with detailed solutions, are included.  These problems come from
several textbooks listed in the reference, in particular from
Folland's {\bfseries Advanced Calculus} \cite{F}.  For many of them,
they are standard problems that one can find in all advanced calculus
textbooks.

The first part of these lecture notes fills the cap that we often
find in undergraduate mathematics programs.  After having learned
calculus of functions of several variables in their first year at the
university, students study basic analysis.  This usually includes some
concepts of metric spaces and topology, the derivative of functions of
several variables including the inverse and implicit function
theorems, and sometime a rigorous introduction to Riemann integration
of functions of one variable.  The rigorous presentation of Riemann
integration for functions of several variables and vector calculus
including integration on manifolds is completely ignored.  The next
time that students may study integration of functions is at the
graduate level when they learn about Lebesgue integrals and so never
cover rigorously vector calculus and integration on manifolds.

The first part of these lecture notes provides a clear, detailed and
complete presentation of Riemann integrals for of functions of several
variables and vector calculus including an introduction to manifolds
and integration on manifolds.  These lecture notes are therefore also ideal
for self study and as a reference for students.

The second part of these lecture notes provides a modern introduction
to differential geometry.  There are several unusual features to the
presentation of this topic in these lecture notes.  One of them is
that the more geometric presentation of the tangent space to a
manifold at a point is used almost everywhere in these notes.  The
only time that the definition of the tangent space of manifold at a
point as a space of linear differential operators is used is to define
the Lie Bracket where it is really the easiest way to define the Lie
Bracket.  Another unusual feature is the inclusion of a chapter on
algebraic geometry.  We feel that this is necessary because algebraic
geometry provides a very nice and efficient technique to reach
conclusions about manifolds that would otherwise require very
technical and convoluted proofs.

The last chapter of these lecture notes is an introduction to
Riemannian geometry.  The presentation of this topic is very geometric
and hopefully provides the reader with a good visual understanding of
the properties associated to Riemannian geometry.  In particular, this
chapter includes a proof of the beautiful Gauss Bonnet theorem.  We
have tried to stay away from formal computations frequently associated
with this subject.

Chapters~\ref{ChaptSingleInt} and \ref{chapMultInt} include a section
with a large number of fully resolved problems.

\section*{Some Terminology}

\subsection*{Metric and Topological Spaces}

The {\bfseries $\mathbf{k}$-norm}\index{Norm!$k$-Norm} of
$\displaystyle \VEC{x} \in \RR^n$ is defined by
\[
\|\VEC{x}\|_k =
\begin{cases}
\displaystyle \left(\sum_{i=1}^n |x_i|^k\right)^{1/k} & \quad \text{if}\ k > 0 \\
\displaystyle \max_{1\leq i \leq n} |x_i| & \quad \text{if} \ k = \infty
\end{cases}
\]
We adopt the convention that $\|\cdot\|$ and $\|\cdot\|_2$ will denote
the {\bfseries $\mathbf{2}$-norm}\index{Norm!$2$-Norm} or
{\bfseries Euclidean norm}\index{Norm!Euclidean Norm}.

The {\bfseries diameter}\index{Diameter} of a set
$\displaystyle W \subset \RR^n$ is the number 
$\diam Q = \inf \{ \|\VEC{x} - \VEC{y}\| : \VEC{x}, \VEC{y} \in W \}$.

The {\bfseries distance}\index{Distance} between two sets
$\displaystyle V, W \subset \RR^n$ is the number
$\dist{V}{W} = \inf \{ \|\VEC{x} - \VEC{y}\| : \VEC{x} \in V \
\text{and} \ \VEC{y} \in W \}$.  The distance between a point
$\displaystyle \VEC{x} \in \RR^n$ and a set
$\displaystyle W \subset \RR^n$ is defined as
$\dist{\VEC{x}}{W} = \dist{\{\VEC{x}\}}{W}$.

The {\bfseries open ball}\index{Open Ball} of radius $r$ centred at
$\displaystyle \VEC{y} \in \RR^n$ is denoted $\displaystyle B_r(\VEC{y})$;
namely,
\[
B_r(\VEC{y}) = \left\{ \VEC{x} \in \RR^n :
  \left\|\VEC{x}-\VEC{y} \right\| < r \right\} \ .
\]
We will also use this notation to denote open balls in other metric spaces.

Let $W$ be a subset in $\displaystyle \RR^n$.  The
{\bfseries closure}\index{Closure of a Set} of $W$, denoted
$\displaystyle \overline{W}$, is the smallest closed set containing
$W$.  We have that $\displaystyle \overline{W} =
\bigcap_{\substack{W\subset V\\V\text{ closed}}} V$.
One of the must important case in these lecture notes is the closure of a
rectangle $\displaystyle R = \prod_{i=1}^n I_i$ where the $I_i$ are
bounded intervals.  We have that
$\displaystyle \overline{R} = \prod_{i=1}^n \overline{I}_i$ where
$\overline{I}_i$ is the closed interval obtained from the interval $I_i$ by
adding the end points.

Let $W$ be a set in $\displaystyle \RR^n$.  The
{\bfseries interior}\index{Interior of a Set} of $W$, denoted
$\displaystyle W^\circ$, is the largest open subset of $W$.  We have that
$\displaystyle W^\circ = \bigcup_{\substack{V\subset W\\V\text{ open}}} V$.
One of the must important case in these lecture notes is the interior of a
rectangle $\displaystyle R = \prod_{i=1}^n I_i$ where the $I_i$ are
bounded intervals.  We have that
$\displaystyle R^\circ = \prod_{i=1}^n I_i^\circ$ where
$\displaystyle I_i^\circ$ is the open interval obtained from the
interval $I_i$ by removing the end points.

The definitions of the interior and closure of a set
$\displaystyle W \subset V \subset \RR^n$ relative to the set $V$ are
identical to the previous definitions except that $\displaystyle \RR^n$
is replaced by $V$ where $V$ has the induced topology from
$\displaystyle \RR^n$.

Let $W$ be a set in $\displaystyle \RR^n$.  The
{\bfseries boundary}\index{Boundary of a Set} of $W$, denoted
$\displaystyle \partial W$, is the set of all points
$\displaystyle \VEC{x} \in \RR^n$ such that
$B_r(\VEC{x}) \cap W \neq \emptyset$
and $\displaystyle B_r(\VEC{x}) \cap (\RR^n \setminus W) \neq \emptyset$
for all $r>0$.
One of the must important case in these lecture notes is the boundary of a
rectangle $\displaystyle R = \prod_{i=1}^n I_i$ where $I_i= [a_i,b_i]$,
$[a_,b_i[$, $]a_i,b_i]$ or $]a_i,b_i[$ for all $i$.  We have that
\begin{align*}
\partial R &= \bigcup_{i=1}^n \Big( \big(
I_1 \times \ldots \times I_{i-1} \times \{ a_i\} \times I_{i+1} \times
\ldots \times I_n \big) \\
&\hspace{7em} \bigcup \big(
I_1 \times \ldots \times I_{i-1} \times \{ b_i\} \times I_{i+1} \times
\ldots \times I_n \big) \Big)
\end{align*}
where $I_1 \times \ldots \times I_{i-1}$ is ignored when $i=1$ and
$I_{i+1} \times \ldots, \times I_n$ is ignored when $i=n$.  The reader
may want to write down the previous formula for the case $n=2$ and
$n=3$ to verify its accuracy.

\subsection*{Function Spaces}

A function $\displaystyle f:[a,b] \to \RR^n$ is
{\bfseries piecewise continuous}\index{Piecewise Continuous Function}
if $f$ is continuous at all but a finite number of points in $[a,b]$
and, at a point $c \in [a,b]$ where $f$ is not continuous, we have that
$\displaystyle \lim_{x\to c^-}f(x)$ and 
$\displaystyle \lim_{x\to c^+}f(x)$ exist.

Similarly, given $m \in \NN$, a function $\displaystyle f:[a,b] \to \RR^n$ is
{\bfseries piecewise $\displaystyle \mathbf{C^m}$ continuously
differentiable}\index{Piecewise $C^m$ Continuous
Differentiable Function} if $f$ is $m$-time differentiable at all but
a finite number of points in $[a,b]$ and, at a point $c \in [a,b]$
where $f$ is not $m$-time differentiable, we have that
$\displaystyle \lim_{x\to c^-}f^{(j)}(x)$ and
$\displaystyle \lim_{x\to c^+}f^{(j)}(x)$ exist for $0 \leq j \leq m$.
Obviously, $f$ is piecewise $\displaystyle C^0$ continuously
differentiable is another way to say that $f$ is
piecewise continuously differentiable.

Let $V$ be a subset of $\displaystyle \RR^n$ and $f:V \to \RR$.  The
{\bfseries supremum norm}\index{Norm!Supremum Norm of a Function}
of $f$ on $V$ is defined by
$\displaystyle \|f\|_{\infty,V} = \sup_{\VEC{x}\in V} |f(\VEC{x})|$.
We will simply write $\|f\|_\infty$ when the set $V$ is clear from the
context.

Let $V$ be a subset of $\displaystyle \RR^n$.
The {\bfseries support}\index{Support of a Function} of a function
$f:V \to \RR$ is the set
$\displaystyle \supp f = \overline{ \{\VEC{x} \in V : f(\VEC{x}) \neq 0 \}}$.

Let $V$ be an open subset of $\displaystyle \RR^n$ and $f:V \to \RR$
be a sufficiently differentiable function, we define the
{\bfseries differential operator
$\displaystyle \mathbf{\diff^{\alpha}}$}\index{Differential Operator} for
$\displaystyle \alpha \in \NN^n$ by
\[
\diff^\alpha f\left(\VEC{x}\right)
= \frac{\partial^{|\alpha|} f}{\partial x_1^{\alpha_1}
\partial x_2^{\alpha_2} \ldots \partial x_n^{\alpha_n}}\left(\VEC{x}\right)
\]
for $\VEC{x} \in V$, where $\displaystyle |\alpha| = \sum_{j=1}^n \alpha_j$.

Let $V$ be an open subset of $\displaystyle \RR^n$.  For $m \in \NN$,
$\displaystyle C^m(V)$ is the space of all
function $f:V \to \RR$ such that all
$\displaystyle \diff^{\alpha}f:V \to \RR$ exists and is
continuous for all $\displaystyle \alpha \in \NN^n$ with
$|\alpha| \leq m$.  Likewise, $\displaystyle C^\infty(V)$ is the space
of all function $f:V \to \RR$ such that all
$\displaystyle \diff^{\alpha}f:V \to \RR$ exists and is
continuous for all $\displaystyle \alpha \in \NN^n$.

For $m\in \NN$, we say that $\displaystyle f \in C^m(\overline{V})$
if $f \in C(\overline{V})$, and there exist and open set
$W \supset \overline{V}$ and a function $\displaystyle F\in C^m(W)$
such that $F = f$ on $V$.  Likewise, we say that
$\displaystyle f \in C^\infty(\overline{V})$ if $f \in C(\overline{V})$,
and there exist and open set $W \supset \overline{V}$ and a function
$\displaystyle F\in C^\infty(W)$ such that $F = f$ on $V$.

Let $\displaystyle f:\RR^n \rightarrow \RR^m$ and
$\displaystyle g:\RR^n \rightarrow \RR$ be two functions.  We write
$f\left(\VEC{x}\right) = O\left(g\left(\VEC{x}\right)\right)$ near
the origin if there exists a positive constant $K$ such that
$\displaystyle \left\| f\left(\VEC{x}\right) \right\| <
K \left| g\left(\VEC{x}\right) \right|$
for $\VEC{x}$ in a neighbourhood of the origin.

\begin{focus*}{Note}
In these lecture notes, $\NN = \{0,1,2,3, \ldots\}$ and
$\displaystyle \NNp = \{1,2,3,\ldots \}$.
\end{focus*}

\begin{focus*}{Note}
By tradition,  $C(V)$ means the space of
continuous real valued functions on $V$.  Namely,
$\displaystyle C(V) = C^0(V)$.  If we really want to talk about the
space of continuously differentiable real valued functions on $V$,
then we will write $\displaystyle C^1(V)$.
\end{focus*}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "notes"
%%% End: 
